---
title: "Correlation metrics with batch characteristics"
author: "Almut Lütge"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  sce_name:
  sce:
  meta:
  metrics:
  summary:
  de:
  last:
  out_cor:
  fig_res:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: hide
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---


```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(CellMixS)
  library(purrr)
  library(tidyr)
  library(plyr)
  library(gridExtra)
  library(scran)
  library(cowplot)
  library(jcolors)
  library(ggpubr)
  library(circlize)
  library(viridis)
  library(ComplexHeatmap)
  library(scran)
  library(magrittr)
  library(colorspace)
    library(corrplot)
    library(RColorBrewer)
    library(dplyr)
  library(hrbrthemes)
})

options(bitmapType='cairo')
```


### Dataset and metrics
```{r data}
sce_name <- unlist(c(strsplit(params$sce_name, ",")))
metrics <- unlist(c(strsplit(params$metrics, ",")))

sce_vec <- c(paste0(params$sce, sce_name, "_", params$last, "_sce.rds"))
names(sce_vec) <- sce_name

meta_vec <- c(paste0(params$meta, sce_name, "_meta.rds"))
names(meta_vec) <- sce_name

sum_vec <- c(paste0(params$summary, sce_name, ".rds"))
names(sum_vec) <- sce_name

de_vec <- c(paste0(params$de, sce_name, ".rds"))
names(de_vec) <- sce_name

out_path_cor <- params$out_cor
out_path_fig <- params$fig_res

cols_data <-c(c(jcolors('pal6'),jcolors('pal8'))[c(1,8,14,5,2:4,6,7,9:13,15:20)],jcolors('pal4'))
names(cols_data) <- c()

cols <-c(c(jcolors('pal6'),jcolors('pal8'), jcolors('pal7'))[c(1,12,18,25,27,2,4,7,3,6,8,14,9,20)],jcolors('pal4'))
names(cols) <- c()
```


## Metric scores and batch characteristics

```{r metric scores, warning=FALSE}
metric_summary <- lapply(sce_name, function(sce_nam, met = metrics){
  sce <- readRDS(sce_vec[sce_nam])
  meta <- readRDS(meta_vec[sce_nam])
  metric_vec <- gsub("_", ".", met)
  metric_vec <- gsub("default", "", metric_vec)
  metric_nam <- unique (grep(paste(metric_vec, collapse="|"),
                             names(colData(sce)), value=TRUE))
  metric_nam <- metric_nam[-grep("smooth", metric_nam)]
  metric_nam <- metric_nam[-grep("casw", metric_nam)]
  #exclude random batches
  if (length(meta[["batch"]]) > 1) {
      batch_str <- paste0("batch", seq(10, 100, 10), collapse="|")
      metric_nam <- metric_nam[-grep(batch_str, metric_nam)]
  }
  # mean score
  d <- colMeans(as.matrix(colData(sce)[, metric_nam]), na.rm = TRUE)
  names(d) <- gsub(paste0('.', meta[["batch"]][1]), '', metric_nam)
  # score variance
  metric_var <- as_tibble(colData(sce)) %>%
      group_by_at(meta[["celltype"]]) %>%
      summarize_at(metric_nam, mean, na.rm = TRUE) %>%
      select(-one_of(meta[["celltype"]])) %>%
      as.matrix() %>% colVars(.,na.rm = TRUE) %>%
      set_names(paste0("ct_var_", names(d)))
  d <- c(d, metric_var)
  d$n_cells <- ncol(sce)
  d$n_batches <- length(levels(as.factor(colData(sce)[,meta$batch[1]])))
  d$n_ct <- length(levels(as.factor(colData(sce)[,meta$celltype])))
  d
}) %>% set_names(sce_name)

metric_summary <- bind_rows(!!!metric_summary) %>% set_rownames(sce_name)

#### ---------- Order by metric type ----------------------------------####### 
#(manual needs to be adjusted if new metrics are added)
metric_fullnam <- colnames(metric_summary)[-grep( "ct_", colnames(metric_summary))]
metric_fullnam <- metric_fullnam[! metric_fullnam %in% c("n_cells", "n_batches", "n_ct")]
cms_ind <- grep("cms", metric_fullnam)
lisi_ind <- grep("isi", metric_fullnam)
ent_ind <- grep("entropy", metric_fullnam)
mm_ind <- grep("mm", metric_fullnam)
asw_ind <- grep("^asw", metric_fullnam)
kbet_ind <- grep("kbet", metric_fullnam)
graph_ind <- grep("graph", metric_fullnam)
pcr_ind <- grep("pcr", metric_fullnam)
metric_order <- metric_fullnam[c(cms_ind[c(2,3,1)], lisi_ind[c(1,3,2)], ent_ind, mm_ind, 
                                       kbet_ind, asw_ind, graph_ind, pcr_ind)]
names(cols) <- metric_fullnam
cols <- cols[metric_fullnam]
####--------------------------------------------------------------------########


```


Per celltype metric scores and batch characteristics
```{r metric scores per celltype}

ct_summary <- lapply(sce_name, function(sce_nam, met = metrics){
    gc()
  sce <- readRDS(sce_vec[sce_nam])
  meta <- readRDS(meta_vec[sce_nam])
  de <- readRDS(de_vec[sce_nam])
  metric_vec <- gsub("_", ".", met)
  metric_vec <- gsub("default", "", metric_vec)
  metric_nam <- unique (grep(paste(metric_vec, collapse="|"),
                             names(colData(sce)), value=TRUE))
  metric_nam <- metric_nam[-grep("smooth", metric_nam)]
  metric_nam <- metric_nam[-grep("casw", metric_nam)]
  #exclude random batches
  if (length(meta[["batch"]]) > 1) {
      batch_str <- paste0("batch", seq(10, 100, 10), collapse="|")
      metric_nam <- metric_nam[-grep(batch_str, metric_nam)]
  }
  nam_met <- gsub(paste0('.', meta[["batch"]][1]), '', metric_nam)

  ct_mean <- as_tibble(colData(sce)) %>%
      group_by_at(meta[["celltype"]]) %>%
      summarize_at(metric_nam, mean, na.rm = TRUE)

  n_de <- lapply(de[["table"]], function(y) vapply(y, function(x) sum(x$PValue < 0.05), numeric(1)))
  n_de_sum <- bind_rows(!!!n_de) %>% colMeans()

  mean_lfc <- lapply(de[["table"]], function(y) vapply(y, function(x) mean(abs(x$logFC)), numeric(1)))
  mean_lfc_sum <- bind_rows(!!!mean_lfc) %>% colMeans()

  mean_lfc_de <- lapply(de[["table"]], function(y) vapply(y, function(x){
      de <- which(x$PValue < 0.05)
      mean(abs(x[de, "logFC"]))
      }, numeric(1)))
  mean_lfc_de_sum <- bind_rows(!!!mean_lfc_de) %>% colMeans()

  ct_mean$n_de <- n_de_sum
  ct_mean$mean_lfc <- mean_lfc_sum
  ct_mean$mean_lfc_de <- mean_lfc_de_sum
  # add number of cells per celltype
  n_ct <- table(colData(sce)[, meta[["celltype"]]])
  ct_mean$n_cells <- n_ct[ct_mean[,meta[["celltype"]]][[1]]]
  ct_mean
}) %>% set_names(sce_name)


```


Summarized batch characteristics
```{r batch summaries}
batch_size <- lapply(sce_name, function(sce_nam){
  sum <- readRDS(sum_vec[sce_nam])
  d <- data.frame("mean_var_batch" = sum$mean_var_batch,
                  "mean_var_celltype" = sum$mean_var_celltype,
                  "mean_de_genes" = sum$mean_mean_n_de_genes,
                  "n_genes" = sum$n_genes_total)
}) %>% set_names(sce_name) %>% bind_rows() %>% set_rownames(sce_name)


size_metric <- cbind(batch_size, metric_summary)

size_metric$batch_origin <- rep(NA, nrow(size_metric))
size_metric[c("pbmc2_media", "pbmc_roche", "csf_media"), "batch_origin"] <- "media"
size_metric[c("csf_patient", "kang", "pbmc2_pat"), "batch_origin"] <- "patient"
size_metric[c("pancreas", "cellbench", "hca"), "batch_origin"] <- "protocol"
size_metric$dataset <- rownames(size_metric)

size_metric <- size_metric %>% arrange(desc(mean_var_batch))
rownames(size_metric) <- size_metric$dataset

#batch type
size_metric <- size_metric %>% mutate("batch_type" = revalue(dataset, c("cellbench" = "linear",
                                                        "csf_media" = "additive",
                                                        "csf_patient" = "additive",
                                                       "hca" = "other",
                                                       "kang" = "other",
                                                       "pancreas" = "interacting",
                                                       "pbmc_roche" = "interacting",
                                                       "pbmc2_media" = "interacting", 
                                                       "pbmc2_pat" = "other")))



```


### Batch size parameter
How does the metric scores changes along size characteristics
```{r size hm}
#colors
q_size <- sequential_hcl(5, palette = "Blues")
col_size <- colorRamp2(c(0.2, 0.1, 0.03, 0.01, 0), q_size)

q_type <- qualitative_hcl(4, palette = "Dark 3")
col_type <- c("protocol" = q_type[1], "media" = q_type[3], "patient" = q_type[4])
col_batch_type <- c("linear" = cols_data[1], "additive" = cols_data[2], 
                    "interacting" = cols_data[3], "other" = cols_data[4])

q_size2 <- sequential_hcl(5, palette = "Mint")
col_size2 <- colorRamp2(c(0.65, 0.5, 0.3, 0.2, 0.1), q_size2)


pal_metr <- "Reds" 
q_metr <- sequential_hcl(5, palette = pal_metr)

col_vec <- function(mean_tab){
    min_col <- min(mean_tab, na.rm = TRUE)
    max_col <- max(mean_tab, na.rm = TRUE)
    mean_col <- mean(mean_tab, na.rm = TRUE)
    q <- quantile(mean_tab, probs = c(0.25, 0.75), na.rm = TRUE)
    col_v <- c(min_col, q[1], mean_col, q[2], max_col)
}


#annotation
ha_type = HeatmapAnnotation("batch_origin" = size_metric$batch_origin,
                            "batch_type" = size_metric$batch_type,
                            simple_anno_size = unit(1, "cm"),
                            col = list("batch_origin" = col_type,
                                       "batch_type" = col_batch_type), 
                            annotation_height = unit(1, "cm"),
                            annotation_name_side = "left")

ha_type_only = HeatmapAnnotation("batch_type" = size_metric$batch_type,
                            simple_anno_size = unit(0.8, "cm"),
                            col = list("batch_type" = col_batch_type), 
                            annotation_height = unit(0.8, "cm"),
                            annotation_name_side = "left")

### Heatmaps

## Size Parameter

# mean variance part
n_batches <- t(as.matrix(size_metric[,c("mean_var_batch", "mean_var_celltype")])) %>% 
  set_colnames(size_metric$dataset)
rownames(n_batches) <- c("var_batch", "var_celltype")

h_size1 <- Heatmap(n_batches, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "variance",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_size,
          top_annotation = ha_type_only,
          rect_gp = gpar(col= "white"))


mean_de_genes <- t(as.matrix(size_metric$mean_de_genes))
colnames(mean_de_genes) <- size_metric$dataset
rownames(mean_de_genes) <- "de_genes"

h_size2 <- Heatmap(mean_de_genes, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "de genes [%]",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_size2,
          rect_gp = gpar(col= "white"))

# Metrics

#scaled matrix
mtr_adjust <- gsub("_", ".", metrics)
mtr_adjust <- gsub(".default", "", mtr_adjust)
metr_str <- paste(mtr_adjust, collapse="|")
metr_columns <- colnames(size_metric)[grep(metr_str, colnames(size_metric))]
metr_columns <- metr_columns[-grep("ct_var", metr_columns)]
scal_mat <- t(apply(size_metric[,metr_columns], 2, function(x){
    scaled <- (x - min(x, na.rm = TRUE))/(max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}))
    
col_scal = colorRamp2(col_vec(scal_mat), q_metr)

h_scal <- Heatmap(scal_mat, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "metrics",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_scal,
          show_heatmap_legend = TRUE,
          rect_gp = gpar(col= "white"))



# cms
cms_columns <- colnames(size_metric)[grep('cms', colnames(size_metric))]
cms_columns <- cms_columns[-grep("ct_var", cms_columns)]
mean_cms <- t(as.matrix(size_metric[, cms_columns]))
colnames(mean_cms) <- size_metric$datasete

col_cms = colorRamp2(col_vec(mean_cms), q_metr)

h_cms <- Heatmap(mean_cms, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "cms",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_cms,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))


# isi's
isi_columns <- colnames(size_metric)[grep('isi', colnames(size_metric))]
isi_columns <- isi_columns[-grep("ct_var", isi_columns)]
mean_isi <- t(as.matrix(size_metric[, isi_columns]))
colnames(mean_isi) <- size_metric$dataset

col_isi = colorRamp2(col_vec(mean_isi), q_metr)

h_isi <- Heatmap(mean_isi, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "isi's",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_isi,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))

# entropy
mean_entropy <- t(as.matrix(size_metric$entropy))
colnames(mean_entropy) <- size_metric$dataset
rownames(mean_entropy) <- "entropy"

col_ent = colorRamp2(col_vec(mean_entropy), q_metr)

h_ent <- Heatmap(mean_entropy, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "entropy",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_ent,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))


# mixingMetric
mean_mm <- t(as.matrix(size_metric$mm))
colnames(mean_mm) <- size_metric$dataset
rownames(mean_mm) <- "mixingMetric"

col_mm = colorRamp2(col_vec(mean_mm), rev(q_metr))

h_mm <- Heatmap(mean_mm, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "mixingMetric",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_mm,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))


# kBet
mean_kbet <- t(as.matrix(size_metric$kbet))
colnames(mean_kbet) <- size_metric$dataset
rownames(mean_kbet) <- "kbet"

col_kbet = colorRamp2(col_vec(mean_kbet), rev(q_metr))

h_kbet <- Heatmap(mean_kbet, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "kbet",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_kbet,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))

# pcr
mean_pcr <- t(as.matrix(size_metric$pcr))
colnames(mean_pcr) <- size_metric$dataset
rownames(mean_pcr) <- "pcr"

col_pcr = colorRamp2(col_vec(mean_pcr), rev(q_metr))

h_pcr <- Heatmap(mean_pcr, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "pcr",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_pcr,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))

# asw
mean_asw <- t(as.matrix(size_metric$asw))
colnames(mean_asw) <- size_metric$dataset
rownames(mean_asw) <- "asw"

col_asw = colorRamp2(col_vec(mean_asw), rev(q_metr))

h_asw <- Heatmap(mean_asw, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "asw",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_asw,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))


# graph_connectivity
mean_gcon <- t(as.matrix(size_metric$graph_connectivity))
colnames(mean_gcon) <- size_metric$dataset
rownames(mean_gcon) <- "graph_connectivity"

col_gcon = colorRamp2(col_vec(mean_gcon), q_metr)

h_gcon <- Heatmap(mean_gcon, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "graph_connectivity",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_gcon,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))

#LEGENDcols
col_fun = colorRamp2(c(0, 0.25, 0.5, 0.75, 1), rev(q_metr))
lgd = Legend(col_fun = col_fun, title = "relative batch size", at = c(0, 0.5, 1), 
    labels = c("low", "median", "high"))


h_list <-  h_size1 %v% h_size2 %v% h_cms %v% h_isi %v% h_mm %v% h_ent %v% h_kbet %v% h_asw %v% h_gcon %v% h_pcr
h_list2 <- h_size1 %v% h_size2 %v% h_scal

draw(h_list, annotation_legend_list = list(lgd))
draw(h_list2)

saveRDS(h_list, paste0(out_path_fig, "_summary_all.rds"))
saveRDS(lgd, paste0(out_path_fig, "_summary_lgd.rds"))
```

Correlation with batch size parameters
```{r size cor, fig.width=10, fig.height=10}
size_columns <- colnames(batch_size)[!colnames(batch_size) %in% "mean_var_celltype"]
size_mat <- size_metric[, c(metr_columns, size_columns)]
size_cor <- cor(size_mat, use = "complete.obs", method = "spearman")

###------------------ metric order -----------------------------------#########
cms_ind <- grep("cms", colnames(size_cor))
lisi_ind <- grep("isi", colnames(size_cor))
ent_ind <- grep("entropy", colnames(size_cor))
mm_ind <- grep("mm", colnames(size_cor))
asw_ind <- grep("^asw", colnames(size_cor))
kbet_ind <- grep("kbet",colnames(size_cor))
graph_ind <- grep("graph", colnames(size_cor))
pcr_ind <- grep("pcr", colnames(size_cor))
size_ind <- grep("mean_", colnames(size_cor))
metric_order <- colnames(size_cor)[c(size_ind, cms_ind[c(2,3,1)], lisi_ind[c(1,3,2)], ent_ind, mm_ind, 
                                       kbet_ind, asw_ind, graph_ind, pcr_ind)]
###------------------ metric reorder -----------------------------------#######
size_cor <- size_cor[metric_order, metric_order]


corrplot(size_cor, 
         type="upper", 
         order="original",
         hclust.method = "complete",
         col=brewer.pal(n=8, name="PuOr"),
         addgrid.col = NA,
         addCoef.col = "black",
         diag = FALSE)

#save correlation
saveRDS(size_cor, out_path_cor)

```

### Score by features {.tabset}
```{r score by feature}

size_met <- size_metric %>% select(!starts_with("ct_"))
size_met <- size_met %>% mutate("colour" = cols_data[as.numeric(as.factor(batch_type))]) 

# get scaled scores
dir_ind <- which(!rownames(scal_mat) %in% c("mm", "kbet", "pcr"))
scal_mat_dir <- scal_mat
scal_mat_dir[dir_ind,] <- 1 - scal_mat[dir_ind,]
size_met_scal <- size_met
size_met_scal[, rownames(scal_mat_dir)] <- t(scal_mat_dir)

feature_vars <- colnames(size_met_scal)[!colnames(size_met_scal) %in% metric_fullnam]
size_long <- size_met_scal %>% pivot_longer(-all_of(feature_vars), names_to = "metric",
                                           values_to = "scaled_score")
size_long$metric <- as.factor(size_long$metric)

sep_plots <- function(feature){
  size_long <- size_long %>% mutate(metric2 = metric)
  cols_rep <- rep(cols[1: length(levels(size_long$metric))], 
                  each = length(levels(as.factor(size_long$dataset))))
  cols_data_rep <- size_long$colour
  size_long$metric <- factor(size_long$metric, levels = metric_order)
  p <- ggplot(size_long, aes_string(x = feature, y = quote(scaled_score))) +
        geom_line(data=size_long %>% dplyr::select(-metric), aes(group=metric2),
                   color="grey", size=0.5, alpha=0.5) + 
        geom_line( aes(color=metric), color=cols_rep, size=1.2 ) + 
        # geom_point( aes(color=batch_type), size=1.4 , color = cols_data_rep) + 
        theme_ipsum(base_family = 'Helvetica') +
        theme(
        legend.position="none",
        plot.title = element_text(size=14),
        panel.grid = element_blank()
      ) +
      ggtitle(feature) +
      facet_wrap(~metric)
  p
}

feature_plot <- feature_vars[!feature_vars %in% c("batch_origin", "batch_type", "dataset", "colour")]  
template_sep <- c(
    "#### {{nm}}\n",
    "```{r scaling sep {{nm}}, echo = FALSE, fig.width = 8, fig.height = 7}\n",
    "p <- sep_plots('{{nm}}')\n",
    "p",
    "saveRDS(p, paste0(out_path_fig, '_{{nm}}.rds'))",
    "```\n",
    "\n"
  )

plots_sep <- lapply(feature_plot, 
  function(nm) knitr::knit_expand(text = template_sep)
)

```

`r knitr::knit(text = unlist(plots_sep))`

### Relation DE and PVEB with score
```{r rel DE score, fig.width=14, fig.height=14}
size_long$metric <- factor(size_long$metric, levels = metric_order)
size_cor_sum <- size_cor[metric_order,c("mean_var_batch", "mean_de_genes")]
size_cor_sum <- size_cor_sum[!rownames(size_cor_sum) %in% c("mean_var_batch", "mean_de_genes"),]

size_long$metric <- recode(size_long$metric, cms.kmin80 = "cms_kmin", cms.bmin80 = "cms_bmin",
       cms = "cms_default", graph_connectivity = "graph", kbet = "kBet")
size_long <- size_long %>% mutate(metric2 = metric)
#transform DE genes to plot both in the same coord system
y1 <- min(size_long$mean_var_batch, na.rm = TRUE)
y2 <- max(size_long$mean_var_batch, na.rm = TRUE)
x1 <- min(size_long$mean_de_genes, na.rm = TRUE)
x2 <- max(size_long$mean_de_genes, na.rm = TRUE)
b <- (y2 - y1) / (x2 - x1)
a <- y1 - b * x1
size_long <- size_long %>% mutate(de_trans = a + b * mean_de_genes)


keep <- c("mean_var_celltype", "mean_de_genes","n_genes","n_cells","n_batches","n_ct","batch_origin","dataset","batch_type", "colour","metric","scaled_score","metric2")
size_long_long <- size_long %>% pivot_longer(-all_of(keep), names_to = "characteristics",
                                           values_to = "variance")

cols_rep <- rep(cols[1:(length(levels(size_long_long$metric))-2)], 
                  each = (2*length(levels(as.factor(size_long_long$dataset)))))
cols_data_rep <- size_long_long$colour


label <- data.frame(
  scaled_score = c(rep(0.25, 10), rep(0.75, 2)), 
  variance = c(rep(0.08, 10), rep(0.06, 2)), 
  metric = levels(size_long$metric)[-c(1:2)],
  label = c(paste0("R_PVE-Batch = ", signif(abs(size_cor_sum[,"mean_var_batch"]), digits = 2),
                   "\nR_DE = ", signif(abs(size_cor_sum[,"mean_de_genes"]), digits = 2)))
)

# Plot
p <- ggplot(size_long_long, aes(x = variance, y = scaled_score)) +
        # geom_line(data=size_long_long %>% dplyr::select(-metric), 
        #           aes(group=metric2), color="grey", 
        #           size=0.5, alpha=0.5) + 
        geom_line( aes(color=metric, linetype=characteristics), 
                   color=cols_rep, size=1.2) +
        scale_x_continuous(sec.axis = sec_axis(as.formula(paste0("~ (. - ", a, "/", b, ")")),
                                               name = 'DE genes')) +
        theme_ipsum(base_family = 'Helvetica') +
        theme(
        plot.title = element_text(size=14),
        panel.grid = element_blank()
      ) +
      ggtitle("Metric performance across datasets") +
      geom_label(data = label, aes(label = label), size = 6) + 
      facet_wrap(~metric)
p

saveRDS(p, paste0(out_path_fig, '_task1.rds'))

```



### Fit a multiple linear regression with batch characteristics and scores {.tabset}

```{r mlr}
size_num <- size_met %>% select_if(is.numeric)
size_vars <- colnames(size_num)[!colnames(size_num) %in% metric_fullnam]

library(jtools)

mlr_fun <- function(met){
  size_tab <- size_num %>% dplyr::select(all_of(c(size_vars, met)))
  f <- as.formula(paste(met, 
        paste(size_vars, collapse = " + "), 
        sep = " ~ "))
  fit <- lm(f, data = size_tab)
  summ(fit, scale = TRUE)
}


template_sep <- c(
    "#### {{nm}}\n",
    "```{r mlr fit {{nm}}, echo = FALSE, fig.width = 8, fig.height = 7}\n",
    "mlr_fun('{{nm}}')\n",
    "```\n",
    "\n"
  )

plots_sep <- lapply(metric_fullnam, 
  function(nm) knitr::knit_expand(text = template_sep)
)

```

`r knitr::knit(text = unlist(plots_sep))`

### Spearman correlation plot
```{r spearman correlation , fig.width=10, fig.height= 6}
size_num_cor <- size_num %>%  as.matrix() %>% cor(., use = "complete.obs", method = "spearman")
feature_rows <- rownames(size_num_cor)[!colnames(size_num_cor) %in% metric_fullnam]
size_num_cor <- size_num_cor[feature_rows, metric_fullnam]

size_num_long <- size_num_cor %>% as_tibble() %>% mutate("feature" = feature_rows) %>% 
  pivot_longer(-feature, names_to = "metric", values_to = "Spearman_correlation")

size_num_long$cor_abs <- abs(size_num_long$Spearman_correlation)
size_num_long$metric <- recode(size_num_long$metric, cms.kmin80 = "cms_kmin", cms.bmin80 = "cms_bmin",
       cms = "cms_default")
metric_order <- c("cms_default", "cms_kmin", "cms_bmin", "lisi", "wisi", "isi", 
                  "entropy", "mm", "asw", "graph_connectivity", "kbet", "pcr")

size_num_long$metric <- factor(size_num_long$metric, levels = metric_order)

p <- ggballoonplot(size_num_long,
              x = 'metric',
              y = 'feature',
              size = 'cor_abs',
              size.range = c(1,22),
              col = "black",
              font.label = c(22, "plain"),
              fill = 'Spearman_correlation') + 
  labs(title="Metric correlation with batch characteristics") +
  theme_ipsum(base_family = 'Helvetica') + 
  guides(size = FALSE) +
  scale_fill_continuous_diverging() +
  theme(axis.text.x = element_text(size=10, angle=45, hjust = 1))

p

saveRDS(p, paste0(out_path_fig, "_spear_cor.rds"))
```

## Relation of DE genes and RVEB with confounding effects (number of cells/genes)
```{r plot confounding, fig.width = 4, fig.height= 12, warning=FALSE}
theme_def <- theme_ipsum(base_family = 'Helvetica',
                    strip_text_face = "bold",
                    axis_title_size = 25,
                    strip_text_size = 22,
                    base_size = 25) +
        theme(
            legend.position="top",
            plot.title = element_text(size=26),
            axis.text.x = element_text(size=16),
            axis.text.y = element_text(size=16),
            panel.grid.major = element_blank(),
            panel.spacing = unit(1, "lines")) 



conf_plot <- function(con_var){
  p1 <- ggplot(size_num, aes_string(x = con_var, y = "mean_var_batch")) +
    geom_point(size = 3, color = "#00AFBB") +
    geom_smooth(method=lm, colour="#00AFBB") +
    theme_def +
    ggtitle(paste0("PVEB"))
  p2 <- ggplot(size_num, aes_string(x = con_var, y = "mean_de_genes")) +
    geom_point(size = 3, color = "#C4961A") +
    geom_smooth(method=lm, colour = "#C4961A") +
    theme_def +
    ggtitle(paste0("mean DE genes"))
  p3 <- ggplot(size_num, aes_string(x = con_var, y = "mean_var_celltype")) +
    geom_point(size = 3, color = "#FC4E07") +
    geom_smooth(method=lm, colour = "#FC4E07") +
    theme_def +
    ggtitle(paste0("PVEC"))
  p_all <- plot_grid(p1,p2,p3, ncol = 1)
  saveRDS(p_all, paste0(out_path_fig, "_cor_rel_vars_", con_var, ".rds"))
  p_all
}

conf_plot("n_cells")
conf_plot("n_genes")
```



## Per celltype metric score

```{r ct metric score}
size_hm <- function(dataset) {
    ### Heatmaps
    ct_mat <- ct_summary[[dataset]]
    ct_mat <- ct_mat[order(ct_mat$mean_lfc_de),]
    n_batches <- t(as.matrix(ct_mat[,c("mean_lfc", "mean_lfc_de")])) 
    col_size <- colorRamp2(rev(col_vec(n_batches)), q_size)
    
    h_size1 <- Heatmap(n_batches, 
          column_title = dataset,
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "mean_lfc",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_size,
          rect_gp = gpar(col= "white"))

    mean_de_genes <- t(as.matrix(ct_mat$n_de))
    rownames(mean_de_genes) <- "de_genes"
    col_size2 <- colorRamp2(rev(col_vec(mean_de_genes)), q_size2)

    h_size2 <- Heatmap(mean_de_genes, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "#de genes",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_size2,
          rect_gp = gpar(col= "white"))
    
    # n_genes
    n_cells <- t(as.matrix(ct_mat$n_cells))
    rownames(n_cells) <- "n_cells"
    q_size3 <- sequential_hcl(5, palette = "Purples")
    col_cells = colorRamp2(col_vec(n_cells), rev(q_size3))
    
    h_size3 <- Heatmap(n_cells, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "#cells",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_cells,
          rect_gp = gpar(col= "white"))

    # cms
    cms_columns <- colnames(ct_mat)[grep('cms', colnames(ct_mat))]
    mean_cms <- t(as.matrix(ct_mat[, cms_columns]))

    col_cms = colorRamp2(col_vec(mean_cms), q_metr)

    h_cms <- Heatmap(mean_cms, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "cms",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_cms,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))

    # isi's
    isi_columns <- colnames(ct_mat)[grep('isi', colnames(ct_mat))]
    mean_isi <- t(as.matrix(ct_mat[, isi_columns]))

    col_isi = colorRamp2(col_vec(mean_isi), q_metr)

    h_isi <- Heatmap(mean_isi, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "isi's",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_isi,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))

    # entropy
    ent_columns <- colnames(ct_mat)[grep('entropy', colnames(ct_mat))]
    mean_entropy <- t(as.matrix(ct_mat[, ent_columns]))
    rownames(mean_entropy) <- "entropy"

    col_ent = colorRamp2(col_vec(mean_entropy), q_metr)

    h_ent <- Heatmap(mean_entropy, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "entropy",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_ent,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))

    # mixingMetric
    mm_columns <- colnames(ct_mat)[grep('mm', colnames(ct_mat))]
    mean_mm <- t(as.matrix(ct_mat[, mm_columns]))
    rownames(mean_mm) <- "mixingMetric"

    col_mm = colorRamp2(col_vec(mean_mm), rev(q_metr))

    h_mm <- Heatmap(mean_mm, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "mixingMetric",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_mm,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))


    # kBet
    kbet_columns <- colnames(ct_mat)[grep('kbet', colnames(ct_mat))]
    mean_kbet <- t(as.matrix(ct_mat[, kbet_columns]))
    rownames(mean_kbet) <- "kbet"
    
    if (min(mean_kbet, na.rm = TRUE) == max(mean_kbet, na.rm = TRUE)) {
        col_kbet = colorRamp2(c(0, 0.25, 0.5, 0.75, 1), rev(q_metr))
    } else {
        col_kbet = colorRamp2(col_vec(mean_kbet), rev(q_metr))
    }
    

    h_kbet <- Heatmap(mean_kbet, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "kbet",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_kbet,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))
    
    # asw
    asw_columns <- colnames(ct_mat)[grep('asw', colnames(ct_mat))]
    mean_asw <- t(as.matrix(ct_mat[, asw_columns]))
    
    if (min(mean_asw, na.rm = TRUE) == max(mean_asw, na.rm = TRUE)) {
        col_asw = colorRamp2(c(0, 0.25, 0.5, 0.75, 1), rev(q_metr))
    } else {
        col_asw = colorRamp2(col_vec(mean_asw), rev(q_metr))
    }

    h_asw <- Heatmap(mean_asw, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "asw",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_asw,
          show_heatmap_legend = FALSE,
          rect_gp = gpar(col= "white"))
    
    #LEGEND
    col_fun = colorRamp2(c(0, 0.25, 0.5, 0.75, 1), rev(q_metr))
    lgd = Legend(col_fun = col_fun, title = "relative batch size", at = c(0, 0.5, 1), 
    labels = c("low", "median", "high"))

    h_list <-  h_size1 %v% h_size2 %v% h_size3 %v% h_cms %v% h_isi %v% h_mm %v% h_ent %v% h_kbet %v% h_asw
    draw(h_list, annotation_legend_list = list(lgd))
}


```



### Celltype score distributions {.tabset}
How does the metric scores change across celltypes
```{r hm, warning=FALSE, message=FALSE, echo=FALSE}

template_hm <- c(
    "#### {{dataset}}\n",
    "```{r hm{{dataset}}, echo = FALSE}\n",
    "h <- size_hm('{{dataset}}')\n",
    "```\n",
    "\n"
  )

plots_hm <- lapply(sce_name, 
  function(dataset) knitr::knit_expand(text = template_hm)
)
```
`r knitr::knit(text = unlist(plots_hm))`

## Celltype metric correlations
Correlation with batch size parameters
```{r size cor ct, fig.width=6, fig.height=6}
cor_fun <- function(dataset){
    ct_mat <- ct_summary[[dataset]]
    # unify metric names
    met <- colnames(ct_mat)
    metric_nam <- met[grep(metr_str, met)]
    ind_kmin <- grep("kmin", metric_nam)
    ind_bmin <- grep("bmin", metric_nam)
    metric_nam <- gsub("_.*", "", metric_nam)
    metric_nam <- gsub("\\..*", "", metric_nam)
    metric_nam[ind_bmin] <- "cms.bmin80"
    metric_nam[ind_kmin] <- "cms.kmin80"
    colnames(ct_mat)[grep(metr_str, met)] <- metric_nam
    # get correlations
    ct_mat <- ct_mat[order(ct_mat$mean_lfc),]
    ct_mat <- ct_mat %>% filter(n_cells > 100) %>% select_if(is.numeric) %>%
        select(-n_cells)
    size_cor <- cor(ct_mat, use = "complete.obs", method = "spearman")
}

cor_list <- lapply(sce_name, cor_fun) %>% set_names(sce_name)
cor_combine <- do.call(rbind, cor_list) 
cor_all <- as.data.frame(cor_combine) %>% 
    mutate("dataset" = rep(names(cor_list), each = nrow(cor_list[[1]])), 
           "metric" =  rownames(cor_combine)) %>% 
    filter(!metric %in% c("mean_lfc", "mean_lfc_de", "n_de"))

#add batch type
cor_all$dataset <- as.factor(cor_all$dataset)
cor_all <- cor_all %>% mutate("batch_type" = revalue(dataset, c("cellbench" = "linear",
                                                        "csf_media" = "additive",
                                                        "csf_patient" = "additive",
                                                       "hca" = "other",
                                                       "kang" = "other",
                                                       "pancreas" = "interacting",
                                                       "pbmc_roche" = "interacting",
                                                       "pbmc2_media" = "interacting", 
                                                       "pbmc2_pat" = "other")))

cor_all[is.na(cor_all)] <- 0

ggplot(cor_all, aes(x = metric , y = mean_lfc_de, fill = batch_type)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(cor_all$metric)))], alpha = 0.5) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols_data) + 
    theme_bw() +
    theme(axis.text.x = element_text(face="bold", size=10, angle=45))


ggplot(cor_all, aes(x = metric , y = mean_lfc, fill = batch_type)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(cor_all$metric)))], alpha = 0.5) +
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols_data) + 
    theme_bw() +
    theme(axis.text.x = element_text(face="bold", size=10, angle=45))


ggplot(cor_all, aes(x = metric , y = n_de, fill = batch_type)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(cor_all$metric)))], alpha = 0.5) +
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols_data) + 
    theme_bw() +
    theme(axis.text.x = element_text(face="bold", size=10, angle=45))


plot_cor <- function(dataset){
    cor_tab <- cor_list[[dataset]]
    cor_tab[is.na(cor_tab)] <- 0
    corrplot(cor_tab, 
         type="upper", 
         order="hclust",
         hclust.method = "complete",
         col=brewer.pal(n=8, name="PuOr"),
         addgrid.col = NA,
         addCoef.col = "black")
}


```


### Celltype score correlations {.tabset}
How does the metric scores correlate across celltypes
```{r cor metric ct, warning=FALSE, message=FALSE, echo=FALSE}
template_cor <- c(
    "#### {{dataset}}\n",
    "```{r cor{{dataset}}, echo = FALSE, fig.width = 9, fig.height = 9}\n",
    "c <- plot_cor('{{dataset}}')\n",
    "```\n",
    "\n"
  )

plots_cor <- lapply(sce_name, 
  function(dataset) knitr::knit_expand(text = template_cor)
)
```

`r knitr::knit(text = unlist(plots_cor))`


```{r}
sessionInfo()
```

