---
title: "metric_scaling_unbalanced"
author: "Almut Lütge"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  sce_name:
  sce:
  metrics:
  last:
  out_cor:
  out_res:
  fig_res:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: hide
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---



Compare different metrics performance on unbalanced datasets.
The strength of the batch effect is equal and just batch proportions are changed.
We test this on different overall batch strength (high, middle, no batch effect).


```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(CellMixS)
  library(purrr)
  library(tidyr)
  library(dplyr)
  library(gridExtra)
  library(scran)
  library(cowplot)
  library(jcolors)
  library(ggpubr)
  library(circlize)
  library(viridis)
  library(ComplexHeatmap)
  library(stringr)
  library(magrittr)
  library(colorspace)
    library(corrplot)
    library(RColorBrewer)
  library(hrbrthemes)
  library(corrplot)
  library(RColorBrewer)
  library(ggforce)
})

options(bitmapType='cairo')
```


### Dataset and metrics
```{r data}
sce_whole_name <- unlist(c(strsplit(params$sce_name, ",")))
sce_name <- gsub("^un_[0-9].*_sim_", "", sce_whole_name) %>% 
    gsub("__[0-9]_sce.*", "", .) %>% unique()

metrics <- unlist(c(strsplit(params$metrics, ",")))

sce_sim_list <- lapply(sce_name, function(dataset){
    sim_vec <- sce_whole_name[grepl(paste0(dataset, "_"), sce_whole_name)]
}) %>% set_names(sce_name)

sim_list <- lapply(sce_name, function(dataset){
    sim_vec <- sce_whole_name[grepl(dataset, sce_whole_name)]
    sim_vec <- gsub("^un_", "", sim_vec) %>% gsub("_sim.*", "", .) %>% 
        as.numeric()
}) %>% set_names(sce_name)


sce_list <- lapply(sce_name, function(dataset){
  sim_vec <- paste0(params$sce, sce_sim_list[[dataset]], "_", params$last, "_sce.rds") %>%
             set_names(sce_sim_list[[dataset]])
}) %>% set_names(sce_name)

out_path_cor <- params$out_cor
out_path_res <- params$out_res
out_path_fig <- params$fig_res

cols_data <-c(c(jcolors('pal6'),jcolors('pal8'))[c(1,8,14,5,2:4,6,7,9:13,15:20)],jcolors('pal4'))
names(cols_data) <- c()

cols <-c(c(jcolors('pal6'),jcolors('pal8'), jcolors('pal7'))[c(12,18,1,25,27,2,4,7,3,6,8,14,9,20)],jcolors('pal4'))
names(cols) <- c()
```

## Metric score trends

### Combined scaled trends {.tabset}

```{r metric scores, warning = FALSE}
metric_vec <- gsub("_", ".", metrics)
metric_vec <- gsub("default", "", metric_vec)

dataset_sim_tab <- function(dataset){
    sim_list <- lapply(sce_sim_list[[dataset]], function(sim_name){
        sce <- readRDS(sce_list[[dataset]][[sim_name]])
        metric_nam <- grep(paste(metric_vec, collapse="|"),
                             names(colData(sce)), value=TRUE)
        metric_nam <- metric_nam[-grep("smooth", metric_nam)]
        res <- as.data.frame(colData(sce)[, metric_nam])
        res$cell <- colnames(sce)
        res
    }) %>% set_names(sce_sim_list[[dataset]])
    sim_tab <- sim_list %>% bind_rows(.id = "dataset")
    metric_vec[metric_vec %in% "cms."] <- "cms.batch_id"
    sim_wide_list <- lapply(metric_vec, function(met){
        met_col <- colnames(sim_tab)[grepl(paste0("^", met), colnames(sim_tab))]
        metr_long <- sim_tab[, c("dataset", "cell", met_col)] 
        metr_long$dataset <- as.factor(metr_long$dataset)
        metr_wide <- pivot_wider(metr_long, names_from = dataset, 
                                 values_from = all_of(met_col))
    }) %>% set_names(metric_vec)
    sim_wide_list
}

res_by_met <- lapply(sce_name, dataset_sim_tab) %>% set_names(sce_name)  

#### ---------- Order by metric type ----------------------------------####### 
#(manual needs to be adjusted if new metrics are added)
cms_ind <- grep("cms", names(res_by_met[[1]]))
lisi_ind <- grep("isi", names(res_by_met[[1]]))
ent_ind <- grep("entropy", names(res_by_met[[1]]))
mm_ind <- grep("mm", names(res_by_met[[1]]))
asw_ind <- grep("sw", names(res_by_met[[1]]))
kbet_ind <- grep("kbet", names(res_by_met[[1]]))
graph_ind <- grep("graph", names(res_by_met[[1]]))
pcr_ind <- grep("pcr", names(res_by_met[[1]]))
metric_order <- names(res_by_met[[1]])[c(cms_ind, lisi_ind, ent_ind, mm_ind, 
                                       kbet_ind, asw_ind, graph_ind, pcr_ind)]
names(cols) <- metric_order
cols <- cols[metric_order]
####--------------------------------------------------------------------########


mean_tab <- lapply(sce_name, function(dataset){
    met_res <- res_by_met[[dataset]]
    metric_vec[metric_vec %in% "cms."] <- "cms.batch_id"
    met_mean <- lapply(metric_vec, function(met){
        mean_vec <- met_res[[met]] %>% select(-cell) %>% as.matrix() %>% colMeans(., na.rm = TRUE)
    }) %>% bind_cols() %>% set_colnames(metric_vec)
    batch_vec <- colnames(met_res[[1]]) %>% str_match("_([0-9].*?)_")
    batch_vec <- as.numeric(batch_vec[!is.na(batch_vec[, 2]), 2])
    met_mean$unbalanced <- batch_vec
    met_mean
}) %>% set_names(sce_name)

## scale means
# scale scores by centering around 0% unbalanced (fully balanced) and 
# scaling to the metrics total range  
scale_means <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    #scaling vector of each method's range
    sce <- readRDS(sce_list[[1]][1])
    n_bids <- length(levels(as.factor(sce$batch_id)))
    mm_ind <- grep("mm", colnames(mean_res))
    lis_ind <- grep("isi", colnames(mean_res))
    scal_vec <- rep(1, ncol(mean_res)-1)
    scal_vec[mm_ind] <- 200
    scal_vec[lis_ind] <- n_bids - 1
    max_res <- mean_res[mean_res$unbalanced == 0, !colnames(mean_res) %in% "unbalanced"]
    scale_res <- mean_res %>% select(-unbalanced) %>%
      scale(., center = max_res, scale = scal_vec) %>% as.data.frame()
    scale_res$unbalanced <- mean_res$unbalanced
    scale_res
}) %>% set_names(names(mean_tab))

#get_descriptive_names
ind_no_batch <- grep("_0$", names(mean_tab))
ind_int_batch <- grep("_0.3$", names(mean_tab))
ind_strong_batch <- grep("_1$", names(mean_tab))
d_names <- names(mean_tab)
d_names[ind_no_batch] <- "no_batch_effect"
d_names[ind_int_batch] <- "moderate_batch_effect"
d_names[ind_strong_batch] <- "separated_batches"
names(scale_means) <- d_names


plot_trends <- function(dataset){
    mean_res <- scale_means[[dataset]]
    mean_long <- mean_res %>% pivot_longer(-unbalanced, names_to = "metric",
                                           values_to = "scaled_result")
    mean_long$metric <- factor(mean_long$metric,levels = metric_order)
    ggplot(mean_long, aes(x = unbalanced, y = scaled_result)) +
        geom_line(aes(color = metric)) + 
        geom_point(aes(color = metric)) +
        theme_bw() +
        scale_color_manual(values = cols) +
        ggtitle(dataset)
}

template <- c(
    "#### {{nm}}\n",
    "```{r scaling{{nm}}, echo = FALSE}\n",
    "plot_trends('{{nm}}')\n",
    "```\n",
    "\n"
  )

plots <- lapply(names(scale_means), 
  function(nm) knitr::knit_expand(text = template)
)
```

`r knitr::knit(text = unlist(plots))`


Plot trends per metric

### Metrics performance separated {.tabset} 

```{r spagetti separated}
# Get lowest level of unbalance that deviates more than 5% from the balanced value. 
limit_tab <- lapply(names(scale_means), function(dataset){
    mean_res <- scale_means[[dataset]]
    dev_lim <- function(x){
        dev_all <- which(abs(x) > 0.05)
        batch <- mean_res$unbalanced[dev_all] %>% min()
    }
    batch_lim <- apply(mean_res, 2, dev_lim)
}) %>% set_names(names(scale_means)) %>% bind_cols() %>% 
  mutate(metric = colnames(scale_means[[1]]))

limit_tab[limit_tab == Inf] <- 1 



sep_trends <- function(dataset){
  mean_res <- scale_means[[dataset]]
  mean_long <- mean_res %>% pivot_longer(-unbalanced, names_to = "metric",
                                           values_to = "scaled_result")
  mean_long <- mean_long %>% mutate(metric2 = metric)
  cols_rep <- rep(cols[1: ncol(mean_res)-1], each = nrow(mean_res))
  mean_long$metric <- factor(mean_long$metric,levels = metric_order)
  lim_ordered <- limit_tab %>% filter(metric %in% metric_order) %>% 
    arrange(match(metric, metric_order))
  find_scal <- function(met){
    lim <- lim_ordered[lim_ordered$metric %in% met, dataset]
    lim_score <- mean_res[mean_res$unbalanced == lim[[1]], met]
  }
  mean_lim <- metric_order %>% map(find_scal) %>% unlist()
  unb_limit <- data.frame("unbalanced" = lim_ordered[,dataset][[1]],
                           "metric" = metric_order,
                           "scaled_score" = mean_lim,
                           "label" = paste0("limit = ", lim_ordered[,dataset][[1]]))
  
  p <- ggplot(mean_long, aes(x = unbalanced, y = scaled_result)) +
        geom_line(data=mean_long %>% dplyr::select(-metric), aes(group=metric2),
                   color="grey", size=0.5, alpha=0.5) + 
        geom_line( aes(color=metric), color=cols_rep, size=1.2 ) + 
        scale_color_viridis(discrete = TRUE) +
        theme_ipsum(base_family = 'Helvetica') +
        theme(
          legend.position="none",
          plot.title = element_text(size=14),
          panel.grid.minor = element_blank()
        ) +
        geom_mark_circle(data = unb_limit, aes(label = label, x = unbalanced, y = scaled_score),
                        label.colour = "black", inherit.aes = FALSE,
                        label.fontsize = 20,
                        size = 1,
                        expand = unit(1, "mm"),
                        label.minwidth = unit(20, "mm")) +
        ggtitle(dataset) +
        facet_wrap(~metric)
  p
}

template_sep <- c(
    "#### {{nm}}\n",
    "```{r scaling sep {{nm}}, echo = FALSE, fig.width = 8, fig.height = 7}\n",
    "p <- sep_trends('{{nm}}')\n",
    "p",
    "saveRDS(p, paste0(out_path_fig, '_{{nm}}.rds'))",
    "```\n",
    "\n"
  )

plots_sep <- lapply(names(scale_means), 
  function(nm) knitr::knit_expand(text = template_sep)
)

```

`r knitr::knit(text = unlist(plots_sep))`


## Plot deviation from mean per metric

### Metrics deviation

```{r metric deviation, warning=FALSE, fig.width=8, fig.height=6}
# Get lowest level of unbalance that deviates more than 5% from the balanced value. 
limit_tab <- lapply(names(scale_means), function(dataset){
    mean_res <- scale_means[[dataset]]
    dev_lim <- function(x){
        dev_all <- which(abs(x) > 0.05)
        batch <- mean_res$unbalanced[dev_all] %>% min()
    }
    batch_lim <- apply(mean_res, 2, dev_lim)
}) %>% set_names(names(scale_means)) %>% bind_cols() %>% 
  mutate(metric = colnames(scale_means[[1]]))

limit_tab[limit_tab == Inf] <- 1 

lim_long <- limit_tab %>% filter(!metric %in% "unbalanced") %>% 
    pivot_longer(-metric, names_to = "dataset", values_to = "unbalanced_limit")

## order by metrics type
lim_long$metric <- factor(lim_long$metric,levels = metric_order)

p <- ggplot(lim_long, aes(x = metric , y = unbalanced_limit, fill = dataset)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(lim_long$metric)))], alpha = 0.5) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols_data) + 
    theme_ipsum(base_family = 'Helvetica') +
    theme(axis.text.x = element_text(face="bold", size=10, angle=45))

p
saveRDS(p, paste0(out_path_fig, "_limits.rds"))

#save mean results
mean_unb_lim <- lim_long %>% group_by(metric) %>% 
  summarise(unb_limit = mean(unbalanced_limit),
            unb_limit_sd = sd(unbalanced_limit))

saveRDS(mean_unb_lim, out_path_res)
```

As lollipop plot
```{r dot plot, fig.width = 10, fig.height=5}
lim_long$metric <- recode(lim_long$metric, cms.kmin = "cms_kmin", cms.bmin = "cms_bmin",
       cms.batch_id = "cms_default", graph.connectivity = "graph", kbet = "kBet")
names(cols) <- c()

p <- ggplot(lim_long) +
  geom_segment( aes(x=dataset, xend=dataset, y=0, yend=unbalanced_limit), color="black") +
  geom_point( aes(x=dataset, y=unbalanced_limit, color=metric, shape = dataset), size=4 ) +
  #coord_flip()+
  theme_ipsum(base_family = 'Helvetica') +
  theme(
    legend.position = "top",
    panel.border = element_blank(),
    panel.spacing = unit(0.1, "lines")
  ) +
  scale_colour_manual(values = c(cols[1:length(levels(as.factor(lim_long$metric)))]), guide=FALSE) +
  xlab("") +
  ylab("Value of Y") +
  facet_wrap(~metric, ncol=length(levels(as.factor(lim_long$metric))), strip.position = "bottom")

p
saveRDS(p, paste0(out_path_fig, "_limits_lolli.rds"))
```



### Correlation between metrics

```{r cor, fig.width=9, fig.height=9}

#scale means
cor_met <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]] %>% select(-unbalanced)
    cor_scal <- cor(mean_res, use = "complete.obs", method = "spearman") 
})

cor_met_mean <- Reduce(`+`, cor_met)/length(cor_met)

corrplot(cor_met_mean, 
         type="upper", 
         order="original",
         hclust.method = "complete",
         col=brewer.pal(n=8, name="PuOr"),
         addgrid.col = NA,
         addCoef.col = "black",
         diag = FALSE)

#save correlation
saveRDS(cor_met_mean, out_path_cor)
```

