---
title: "Metrics Scaling simulation"
author: "Almut Lütge"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  sce_name:
  sce:
  meta:
  last:
  metrics:
  out_cor:
  out_res:
  fig_res:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: show
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---


[Back to home](index.html)

```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(scater)
  library(CellMixS)
  library(purrr)
  library(tidyr)
  library(dplyr)
  library(gridExtra)
  library(scran)
  library(cowplot)
  library(jcolors)
  library(ggpubr)
  library(stringr)
  library(viridis)
  library(scales)
  library(hrbrthemes)
  library(magrittr)
  library(forcats)
  library(corrplot)
  library(RColorBrewer)
  library(ggforce)
})

options(bitmapType='cairo')
```


### Dataset and metrics
```{r data}
last <- params$last
sce_whole_name <- unlist(c(strsplit(params$sce_name, ",")))
sce_name <- gsub("^sim_", "", sce_whole_name) %>% 
    gsub("_[0-9].*", "", .) %>% unique()

metrics <- unlist(c(strsplit(params$metrics, ",")))

sce_sim_list <- lapply(sce_name, function(dataset){
    sim_vec <- sce_whole_name[grepl(dataset, sce_whole_name)]
}) %>% set_names(sce_name)


sce_list <- lapply(sce_name, function(dataset){
  sim_vec <- paste0(params$sce, sce_sim_list[[dataset]], "_", last, "_sce.rds") %>%
             set_names(sce_sim_list[[dataset]])
}) %>% set_names(sce_name)


meta_list <- lapply(sce_name, function(dataset){
  sim_vec <- paste0(params$meta, sce_sim_list[[dataset]], "_meta.rds") %>%
             set_names(sce_sim_list[[dataset]])
}) %>% set_names(sce_name)


out_path_cor <- params$out_cor
out_path_res <- params$out_res
out_path_fig <- params$fig_res

cols_data <-c(c(jcolors('pal6'),jcolors('pal8'))[c(1,8,14,5,2:4,6,7,9:13,15:20)],jcolors('pal4'))
names(cols_data) <- c()

cols <-c(c(jcolors('pal6'),jcolors('pal8'), jcolors('pal7'))[c(1,12,18,25,27,2,4,7,3,6,8,14,9,20)],jcolors('pal4'))
names(cols) <- c()

```

## Scaling with simulated batches

### Overall scaling {.tabset}
How do the metric scores scale with simulated batch effects that are gradually increased?
```{r vi hist, warning=FALSE, message=FALSE}

metric_vec <- gsub("_", ".", metrics)
metric_vec <- gsub("default", "", metric_vec)

dataset_sim_tab <- function(dataset){
    sim_list <- lapply(sce_sim_list[[dataset]], function(sim_name){
        sce <- readRDS(sce_list[[dataset]][[sim_name]])
        meta <- readRDS(meta_list[[dataset]][[sim_name]])
        metric_nam <- grep(paste(metric_vec, collapse="|"),
                             names(colData(sce)), value=TRUE)
        metric_nam <- metric_nam[-grep("smooth", metric_nam)]
        res <- as.data.frame(colData(sce)[, metric_nam])
        res$cell <- colnames(sce)
        res
    }) %>% set_names(sce_sim_list[[dataset]])
    sim_tab <- sim_list %>% bind_rows(.id = "dataset")
    metric_vec[metric_vec %in% "cms."] <- "cms.batch_id"
    sim_wide_list <- lapply(metric_vec, function(met){
        met_col <- colnames(sim_tab)[grepl(paste0("^", met), colnames(sim_tab))]
        metr_long <- sim_tab[, c("dataset", "cell", met_col)] 
        metr_long$dataset <- as.factor(metr_long$dataset)
        metr_wide <- pivot_wider(metr_long, names_from = dataset, 
                                 values_from = met_col)
    }) %>% set_names(metric_vec)
}

res_by_met <- lapply(sce_name, dataset_sim_tab) %>% set_names(sce_name)

#### ---------- Order by metric type ----------------------------------####### 
#(manual needs to be adjusted if new metrics are added)
cms_ind <- grep("cms", names(res_by_met[[1]]))
lisi_ind <- grep("isi", names(res_by_met[[1]]))
ent_ind <- grep("entropy", names(res_by_met[[1]]))
mm_ind <- grep("mm", names(res_by_met[[1]]))
asw_ind <- grep("sw", names(res_by_met[[1]]))
kbet_ind <- grep("kbet", names(res_by_met[[1]]))
graph_ind <- grep("graph", names(res_by_met[[1]]))
pcr_ind <- grep("pcr", names(res_by_met[[1]]))
metric_order <- names(res_by_met[[1]])[c(cms_ind, lisi_ind, ent_ind, mm_ind, 
                                       kbet_ind, asw_ind, graph_ind, pcr_ind)]

names(cols) <- names(res_by_met[[1]])
#cols <- cols[names(res_by_met[[1]])]
####--------------------------------------------------------------------########

mean_tab <- lapply(sce_name, function(dataset){
    met_res <- res_by_met[[dataset]]
    metric_vec[metric_vec %in% "cms."] <- "cms.batch_id"
    met_mean <- lapply(metric_vec, function(met){
        mean_vec <- met_res[[met]] %>% select(-cell) %>% as.matrix() %>% colMeans(., na.rm = TRUE)
    }) %>% bind_cols() %>% set_colnames(metric_vec)
    batch_vec <- colnames(met_res[[1]]) %>% str_match("_([0-9].*?)_")
    batch_vec <- as.numeric(batch_vec[!is.na(batch_vec[, 2]), 2])
    met_mean$batch <- batch_vec
    met_mean
}) %>% set_names(sce_name)

#scale means
scale_means <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    dir_ind <- which(!colnames(mean_res) %in% c("batch", "mm", "kbet", "pcr"))
    mean_res[, dir_ind] <- mean_res[, dir_ind] * -1
    scale_res <- mean_res %>% select(-batch) %>% scale() %>% as.data.frame()
    scale_res$batch <- mean_res$batch
    scale_res
}) %>% set_names(names(mean_tab))


plot_trends <- function(dataset){
    mean_res <- scale_means[[dataset]]
    mean_long <- mean_res %>% pivot_longer(-batch, names_to = "metric",
                                           values_to = "scaled_score")
    mean_long$metric <- factor(mean_long$metric,levels = metric_order)
    ggplot(mean_long, aes(x = batch, y = scaled_score)) +
        geom_line(aes(color = metric)) + 
        geom_point(aes(color = metric)) +
        theme_bw() +
        scale_color_manual(values = cols) +
        ggtitle(dataset)
}

template <- c(
    "#### {{nm}}\n",
    "```{r scaling{{nm}}, echo = FALSE}\n",
    "plot_trends('{{nm}}')\n",
    "```\n",
    "\n"
  )

plots <- lapply(names(scale_means), 
  function(nm) knitr::knit_expand(text = template)
)
```

`r knitr::knit(text = unlist(plots))`



### Compare detection limits and correlation within limits {.tabset}
To compare the metrics limits and their relative correlation with batch 
strength/size we define an upper limit at 95% of the maximal metrics score and 
a minimal limit at the minimal metrics score and use the corresponding 
batch size factor as batch limits. We check the metrics correlation with the 
batch size factor between it's limits. 
Be aware that these limits are dataset specific and do not represent fixed 
limits. They can only be used to compare metrics, but are not informative about 
the absolut metric's sensitivity.

```{r limits, warning=FALSE, message=FALSE }

limit_list <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    lim_rev <- function(x){
        percent <- (max(x) - min(x))/100
        lower <- which(x < (max(x) - 10 * percent)) # smallest value above 0.95 * batch0
        upper <- which(x > (min(x) + 10 * percent))
        upper_lim <- which(x < (min(x) + 1 * percent))
        lower_lim <- which(x > (max(x) - 1 * percent))
        min_bat <- min(mean_res$batch[which(x == max(x))])
        max_bat <- min(mean_res$batch[which(x == min(x))])
        res_low <- min(mean_res$batch[lower][mean_res$batch[lower] > min_bat])
        res_up <- min(mean_res$batch[upper][mean_res$batch[upper] > max_bat])
        res_up_limit <-  min(mean_res$batch[upper_lim])
        res_low_limit <-  min(mean_res$batch[lower_lim])
        batch <- c(res_low, res_up, res_up_limit, res_low_limit)
    }
    batch_lim <- apply(mean_res, 2, lim_rev) %>% set_rownames(c("lower", "upper", "limit", "limit2"))
    batch_lim[,c("kbet", "mm", "pcr")] <- 
      batch_lim[c(2,1,4,3), c("kbet", "mm", "pcr")]
    batch_lim
}) %>% set_names(names(mean_tab))

lim_fun <- function(dataset){
    lim_res <- limit_list[[dataset]]
    lim_low <- lim_res["lower",] %>% set_names(colnames(lim_res))
}

limit_tab <- names(limit_list) %>% map(lim_fun) %>% 
    set_names(names(limit_list)) %>% bind_cols() %>% 
    mutate(metric = colnames(limit_list[[1]]))


lim_long <- limit_tab %>% filter(!metric %in% "batch") %>% 
    pivot_longer(-metric, names_to = "dataset", values_to = "batch_limit")


cor_list <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    limit_res <- limit_list[[dataset]]
    cor_fun <- function(metric){
        cor_res <- mean_res  %>%
            filter(batch <= limit_res["limit", metric]) %>%
            select(metric, "batch") %>% cor(method ="pearson") %>% min() %>% abs()
        #%>% filter(batch >= limit_res["lower", metric])
    }
    cor_tab <- map(colnames(mean_res), cor_fun) %>% bind_cols() %>% 
        set_colnames(colnames(mean_res)) 
}) %>% bind_rows() %>% set_rownames(names(mean_tab))

cor_long <- cor_list  %>% mutate(dataset = rownames(.)) %>% select(-batch) %>% 
    pivot_longer(-dataset, names_to = "metric", values_to = "correlation")


cor_list_all <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    limit_res <- limit_list[[dataset]]
    cor_fun <- function(metric){
        cor_res <- mean_res  %>%
            select(metric, "batch") %>% cor(method ="spearman") %>% min() %>% abs()
    }
    cor_tab <- map(colnames(mean_res), cor_fun) %>% bind_cols() %>% 
        set_colnames(colnames(mean_res)) 
}) %>% bind_rows() %>% set_rownames(names(mean_tab))

cor_long_all <- cor_list_all  %>% mutate(dataset = rownames(.)) %>% select(-batch) %>% 
    pivot_longer(-dataset, names_to = "metric", values_to = "correlation")


### Mutual information
# mut_list <- lapply(names(mean_tab), function(dataset){
#     mean_res <-  mean_tab[[dataset]]
#     mut <- cmi(mean_res)
#     norm_mi <- mut$mi[ncol(mut$mi), nrow(mut$mi)]
#     norm_vec <- rep(sqrt(norm_mi * norm_mi), ncol(mean_res))
#     norm_mut <- mut$mi[,ncol(mut$mi)]/norm_vec %>% set_names(colnames(mean_res))
# }) %>% bind_rows() %>% set_rownames(names(mean_tab))
# 
# mut_long <- mut_list  %>% mutate(dataset = rownames(.)) %>% select(-batch) %>% 
#     pivot_longer(-dataset, names_to = "metric", values_to = "MI")

# # order by metrics type
lim_long$metric <- factor(lim_long$metric,levels = metric_order)
cor_long$metric <- factor(cor_long$metric,levels = metric_order)
cor_long_all$metric <- factor(cor_long_all$metric,levels = metric_order)
#mut_long$metric <- factor(mut_long$metric,levels = metric_order)
```


#### Batch limits
```{r batch lims}
p <- ggplot(lim_long, aes(x = metric , y = batch_limit, fill = metric)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(lim_long$metric)))], alpha = 0.5) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols) + 
    theme_ipsum(base_family = 'Helvetica') +
    theme(axis.text.x = element_text(size=10, angle=45, hjust = 1))

p
saveRDS(p, paste0(out_path_fig, "_limits.rds"))
```

#### Relative strength correlation
```{r strength correlation}
p <- ggplot(cor_long, aes(x = metric , y = correlation, fill = metric)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(cor_long$metric)))], alpha = 0.5) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols) + 
    theme_ipsum(base_family = 'Helvetica') +
    theme(axis.text.x = element_text(size=10, angle=45, hjust = 1))

p
saveRDS(p, paste0(out_path_fig, "_rel_cor.rds"))
```

#### Total strength correlation
```{r strength correlation total}
p <- ggplot(cor_long_all, aes(x = metric , y = correlation, fill = metric)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(cor_long$metric)))], alpha = 0.5) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols) + 
    theme_ipsum(base_family = 'Helvetica') +
    theme(axis.text.x = element_text(size=10, angle=45, hjust = 1))

p
saveRDS(p, paste0(out_path_fig, "_cor.rds"))
```

<!-- #### Mutual information -->
<!-- ```{r mutual information} -->
<!-- p <- ggplot(mut_long, aes(x = metric , y = MI, fill = metric)) +  -->
<!--     geom_boxplot(fill = cols[1:length(levels(as.factor(cor_long$metric)))], alpha = 0.5) +  -->
<!--     geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) + -->
<!--     scale_fill_manual(values=cols) +  -->
<!--     theme_ipsum(base_family = 'Helvetica') + -->
<!--     theme(axis.text.x = element_text(size=10, angle=45, hjust = 1)) -->

<!-- p -->

<!-- ``` -->

#### Plot trade-off between strength correlation and detection limit
```{r relationship}
#join datasets
trade_long <- full_join(cor_long_all, lim_long, by = c("dataset", "metric"))
trade_sorted <- trade_long %>% group_by(metric) %>% 
  summarise(mean_correlation = mean(correlation),
            mean_limit = mean(batch_limit),
            correlation_sd = sd(correlation),
            batch_limit_sd = sd(batch_limit))


lev_metric_level <- c(cell = c("cms.bmin", "cms.batch_id","cms.kmin",
                               "isi", "wisi", "lisi", "entropy", "mm"),
                      celltype = c("kbet", "asw", "graph.connectivity"),
                      global = "pcr")
  
trade_sorted$metric_level <- fct_collapse(factor(trade_sorted$metric), 
                                          cell = c("cms.bmin", "cms.batch_id", 
                                                   "cms.kmin", "isi", "wisi", 
                                                   "lisi", "entropy", "mm"), 
                                          celltype = c("kbet", "asw", "sw", 
                                                       "casw", "graph.connectivity"), 
                                          global = "pcr")

ggplot(trade_sorted, aes(x=mean_limit, y=mean_correlation, color=metric)) +
    geom_pointrange(aes(ymin=mean_correlation - correlation_sd, 
                        ymax=mean_correlation + correlation_sd)) +
    geom_pointrange(aes(xmin=mean_limit - batch_limit_sd, 
                        xmax=mean_limit + batch_limit_sd)) +
    scale_colour_manual(values=cols) +
    ggtitle("Trade off: batch detection limit - correlation") +
    theme_ipsum(base_family = 'Helvetica')

```

#### Trade off top metrics
```{r trade off top metrics}
trade_top <- trade_sorted %>% filter(mean_limit < 0.6, mean_correlation > 0.7)

ggplot(trade_top, aes(x=mean_limit, y=mean_correlation, color=metric)) +
    geom_pointrange(aes(ymin=mean_correlation - correlation_sd, 
                        ymax=mean_correlation + correlation_sd)) +
    geom_pointrange(aes(xmin=mean_limit - batch_limit_sd, 
                        xmax=mean_limit + batch_limit_sd)) +
    scale_colour_manual(values = cols) +
    ggtitle("Trade off: batch detection limit - correlation") +
    theme_ipsum(base_family = 'Helvetica')

```


#### Trade-off mean_only
```{r relationship points, fig.height=5, fig.width=8.5}
#join datasets
p <- ggplot(trade_sorted, aes(x=mean_limit, y=mean_correlation, color=metric, shape = metric_level)) +
    geom_point(size = 8, color = "black", ) +
    geom_point(size = 7) +
    scale_colour_manual(values=cols) +
    ggtitle("Trade off: Sensitivity - scaling") +
    theme_ipsum(base_family = 'Helvetica') +
    theme(legend.box = "vertical")

p
saveRDS(p, paste0(out_path_fig, "_trade_off.rds"))
```
#### Trade-off using relative correlation
```{r rel corr, fig.height=5, fig.width=8.5}
#join datasets
p <- ggplot(trade_sorted, aes(x=mean_limit, y=mean_correlation, color=metric, shape = metric_level)) +
    geom_point(size = 8, color = "black", ) +
    geom_point(size = 7) +
    scale_colour_manual(values=cols) +
    ggtitle("Trade off: Sensitivity - scaling") +
    theme_ipsum(base_family = 'Helvetica') +
    theme(legend.box = "vertical")

p
saveRDS(p, paste0(out_path_fig, "_trade_off_relative.rds"))
```

### Correlation between metrics

Check mean correlation
```{r cor, fig.width=9, fig.height=9}

#scale means
scale_m <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    scale_res <- mean_res %>% select(-batch) %>% scale() %>% as.data.frame()
}) %>% set_names(names(mean_tab))

cor_met <- lapply(names(scale_m), function(dataset){
    scal_res <- scale_m[[dataset]]
    cor_scal <- cor(scal_res, use = "complete.obs", method = "spearman") 
})

cor_met_mean <- Reduce(`+`, cor_met)/length(cor_met)

corrplot(cor_met_mean, 
         type="upper", 
         order="original",
         hclust.method = "complete",
         col=brewer.pal(n=8, name="PuOr"),
         addgrid.col = NA,
         addCoef.col = "black",
         diag = FALSE)

#save correlation
saveRDS(cor_met_mean, out_path_cor)
saveRDS(trade_sorted, out_path_res)
```


### Metrics performance separated {.tabset} 

```{r spagetti separated, fig.width = 8, fig.height = 7}

sep_trends <- function(dataset){
  mean_res <- scale_means[[dataset]]
  dat_nam <- dataset
  cor_dat <- cor_long %>% dplyr::filter(dataset %in% dat_nam) %>% 
    arrange(match(metric, metric_order)) %>% select(correlation) %>% 
    abs() %>% signif(digits = 2)
  label <- data.frame(
    scaled_score = c(rep(-1, 12)), 
    batch = c(rep(2.5, 12)), 
    metric = metric_order,
    label = c(paste0("R = ", cor_dat$correlation))
  )
  lim_ordered <- limit_tab %>% dplyr::filter(metric %in% metric_order) %>%
    arrange(match(metric, metric_order))
  find_scal <- function(met){
    lim <- lim_ordered[lim_ordered$metric %in% met, dataset]
    lim_score <- mean_res[mean_res$batch == lim[[1]], met]
  }
  mean_lim <- metric_order %>% map(find_scal) %>% unlist()
  sens_limit <- data.frame("batch" = lim_ordered[,dataset][[1]],
                           "metric" = metric_order,
                           "scaled_score" = mean_lim,
                           "label" = paste0("limit = ", lim_ordered[,dataset][[1]]))
  mean_long <- mean_res %>% pivot_longer(-batch, names_to = "metric",
                                           values_to = "scaled_score")
  mean_long <- mean_long %>% mutate(metric2 = metric)
  cols_rep <- rep(cols[1: ncol(mean_res)-1], each = nrow(mean_res))
  mean_long$metric <- factor(mean_long$metric,levels = metric_order)
  p <- ggplot(mean_long, aes(x = batch, y = scaled_score)) +
        geom_line(data=mean_long %>% dplyr::select(-metric), aes(group=metric2),
                   color="grey", size=0.5, alpha=0.5) + 
        geom_line( aes(color=metric), color=cols_rep, size=1.2 ) + 
        theme_ipsum(base_family = 'Helvetica') +
        theme(
        legend.position="none",
        plot.title = element_text(size=14),
        panel.grid = element_blank()
      ) +
      geom_label(data = label, aes(label = label), size = 6) + 
      geom_mark_circle(data = sens_limit, aes(label = label, x = batch, y = scaled_score),
                        label.colour = "black", inherit.aes = FALSE,
                        label.fontsize = 16,
                        size = 1,
                        expand = unit(1, "mm"),
                        label.minwidth = unit(50, "mm")) +
      ggtitle(dataset) +
      facet_wrap(~metric)
    p
}

template_sep <- c(
    "#### {{nm}}\n",
    "```{r scaling sep {{nm}}, echo = FALSE, fig.width = 8, fig.height = 7}\n",
    "p <- sep_trends('{{nm}}')\n",
    "p",
    "saveRDS(p, paste0(out_path_fig, '_{{nm}}.rds'))",
    "```\n",
    "\n"
  )

plots_sep <- lapply(names(scale_means), 
  function(nm) knitr::knit_expand(text = template_sep)
)

```

`r knitr::knit(text = unlist(plots_sep))`




