---
title: "Metrics Scaling simulation"
author: "Almut Lütge"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  sce_name:
  sce:
  meta:
  last:
  metrics:
  out_cor:
  out_res:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: show
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---


[Back to home](index.html)

```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(scater)
  library(CellMixS)
  library(purrr)
  library(tidyr)
  library(dplyr)
  library(gridExtra)
  library(scran)
  library(cowplot)
  library(jcolors)
  library(ggpubr)
  library(stringr)
  library(viridis)
  library(scales)
  library(hrbrthemes)
  library(magrittr)
  library(forcats)
  library(corrplot)
  library(RColorBrewer)
})

options(bitmapType='cairo')
```


### Dataset and metrics
```{r data}
last <- params$last
sce_whole_name <- unlist(c(strsplit(params$sce_name, ",")))
sce_name <- gsub("^sim_", "", sce_whole_name) %>% 
    gsub("_[0-9].*", "", .) %>% unique()

metrics <- unlist(c(strsplit(params$metrics, ",")))

sce_sim_list <- lapply(sce_name, function(dataset){
    sim_vec <- sce_whole_name[grepl(dataset, sce_whole_name)]
}) %>% set_names(sce_name)


sce_list <- lapply(sce_name, function(dataset){
  sim_vec <- paste0(params$sce, sce_sim_list[[dataset]], "_", last, "_sce.rds") %>%
             set_names(sce_sim_list[[dataset]])
}) %>% set_names(sce_name)


meta_list <- lapply(sce_name, function(dataset){
  sim_vec <- paste0(params$meta, sce_sim_list[[dataset]], "_meta.rds") %>%
             set_names(sce_sim_list[[dataset]])
}) %>% set_names(sce_name)


out_path_cor <- params$out_cor
out_path_res <- params$out_res

cols_data <-c(c(jcolors('pal6'),jcolors('pal8'))[c(1,8,14,5,2:4,6,7,9:13,15:20)],jcolors('pal4'))
names(cols_data) <- c()

cols <-c(c(jcolors('pal6'),jcolors('pal8'), jcolors('pal7'))[c(1,12,18,25,27,2,4,7,3,6,8,14,9,20)],jcolors('pal4'))
names(cols) <- c()

```

## Scaling with simulated batches

### Overall scaling {.tabset}
How do the metric scores scale with simulated batch effects that are gradually increased?
```{r vi hist, warning=FALSE, message=FALSE}

metric_vec <- gsub("_", ".", metrics)
metric_vec <- gsub("default", "", metric_vec)

dataset_sim_tab <- function(dataset){
    sim_list <- lapply(sce_sim_list[[dataset]], function(sim_name){
        sce <- readRDS(sce_list[[dataset]][[sim_name]])
        meta <- readRDS(meta_list[[dataset]][[sim_name]])
        metric_nam <- grep(paste(metric_vec, collapse="|"),
                             names(colData(sce)), value=TRUE)
        metric_nam <- metric_nam[-grep("smooth", metric_nam)]
        res <- as.data.frame(colData(sce)[, metric_nam])
        res$cell <- colnames(sce)
        res
    }) %>% set_names(sce_sim_list[[dataset]])
    sim_tab <- sim_list %>% bind_rows(.id = "dataset")
    metric_vec[metric_vec %in% "cms."] <- "cms.batch_id"
    sim_wide_list <- lapply(metric_vec, function(met){
        met_col <- colnames(sim_tab)[grepl(paste0("^", met), colnames(sim_tab))]
        metr_long <- sim_tab[, c("dataset", "cell", met_col)] 
        metr_long$dataset <- as.factor(metr_long$dataset)
        metr_wide <- pivot_wider(metr_long, names_from = dataset, 
                                 values_from = met_col)
    }) %>% set_names(metric_vec)
}

res_by_met <- lapply(sce_name, dataset_sim_tab) %>% set_names(sce_name)

#### ---------- Order by metric type ----------------------------------####### 
#(manual needs to be adjusted if new metrics are added)
cms_ind <- grep("cms", names(res_by_met[[1]]))
lisi_ind <- grep("isi", names(res_by_met[[1]]))
ent_ind <- grep("entropy", names(res_by_met[[1]]))
mm_ind <- grep("mm", names(res_by_met[[1]]))
asw_ind <- grep("sw", names(res_by_met[[1]]))
kbet_ind <- grep("kbet", names(res_by_met[[1]]))
graph_ind <- grep("graph", names(res_by_met[[1]]))
pcr_ind <- grep("pcr", names(res_by_met[[1]]))
metric_order <- names(res_by_met[[1]])[c(cms_ind, lisi_ind, ent_ind, mm_ind, 
                                       kbet_ind, asw_ind, graph_ind, pcr_ind)]
names(cols) <- names(res_by_met[[1]])
#cols <- cols[names(res_by_met[[1]])]
####--------------------------------------------------------------------########

mean_tab <- lapply(sce_name, function(dataset){
    met_res <- res_by_met[[dataset]]
    metric_vec[metric_vec %in% "cms."] <- "cms.batch_id"
    met_mean <- lapply(metric_vec, function(met){
        mean_vec <- met_res[[met]] %>% select(-cell) %>% as.matrix() %>% colMeans(., na.rm = TRUE)
    }) %>% bind_cols() %>% set_colnames(metric_vec)
    batch_vec <- colnames(met_res[[1]]) %>% str_match("_([0-9].*?)_")
    batch_vec <- as.numeric(batch_vec[!is.na(batch_vec[, 2]), 2])
    met_mean$batch <- batch_vec
    met_mean
}) %>% set_names(sce_name)

#scale means
scale_means <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    dir_ind <- which(!colnames(mean_res) %in% c("batch", "mm", "kbet", "pcr",
                                                "graph.connectivity"))
    mean_res[, dir_ind] <- mean_res[, dir_ind] * -1
    scale_res <- mean_res %>% select(-batch) %>% scale() %>% as.data.frame()
    scale_res$batch <- mean_res$batch
    scale_res
}) %>% set_names(names(mean_tab))


plot_trends <- function(dataset){
    mean_res <- scale_means[[dataset]]
    mean_long <- mean_res %>% pivot_longer(-batch, names_to = "metric",
                                           values_to = "scaled_score")
    mean_long$metric <- factor(mean_long$metric,levels = metric_order)
    ggplot(mean_long, aes(x = batch, y = scaled_score)) +
        geom_line(aes(color = metric)) + 
        geom_point(aes(color = metric)) +
        theme_bw() +
        scale_color_manual(values = cols) +
        ggtitle(dataset)
}

template <- c(
    "#### {{nm}}\n",
    "```{r scaling{{nm}}, echo = FALSE}\n",
    "plot_trends('{{nm}}')\n",
    "```\n",
    "\n"
  )

plots <- lapply(names(scale_means), 
  function(nm) knitr::knit_expand(text = template)
)
```

`r knitr::knit(text = unlist(plots))`


### Metrics performance separated {.tabset} 

```{r spagetti separated, fig.width = 8, fig.height = 7}

sep_trends <- function(dataset){
  mean_res <- scale_means[[dataset]]
  mean_long <- mean_res %>% pivot_longer(-batch, names_to = "metric",
                                           values_to = "scaled_score")
  mean_long <- mean_long %>% mutate(metric2 = metric)
  cols_rep <- rep(cols[1: ncol(mean_res)-1], each = nrow(mean_res))
  mean_long$metric <- factor(mean_long$metric,levels = metric_order)
  p <- ggplot(mean_long, aes(x = batch, y = scaled_score)) +
        geom_line(data=mean_long %>% dplyr::select(-metric), aes(group=metric2),
                   color="grey", size=0.5, alpha=0.5) + 
        geom_line( aes(color=metric), color=cols_rep, size=1.2 ) + 
        theme_ipsum() +
        theme(
        legend.position="none",
        plot.title = element_text(size=14),
        panel.grid = element_blank()
      ) +
      ggtitle(dataset) +
      facet_wrap(~metric)
    p
}

template_sep <- c(
    "#### {{nm}}\n",
    "```{r scaling sep {{nm}}, echo = FALSE, fig.width = 8, fig.height = 7}\n",
    "sep_trends('{{nm}}')\n",
    "```\n",
    "\n"
  )

plots_sep <- lapply(names(scale_means), 
  function(nm) knitr::knit_expand(text = template_sep)
)

```

`r knitr::knit(text = unlist(plots_sep))`


### Compare detection limits and correlation within limits {.tabset}
To compare the metrics limits and their relative correlation with batch 
strength/size we define an upper limit at 95% of the maximal metrics score and 
a minimal limit at the minimal metrics score and use the corresponding 
batch size factor as batch limits. We check the metrics correlation with the 
batch size factor between it's limits. 
Be aware that these limits are dataset specific and do not represent fixed 
limits. They can only be used to compare metrics, but are not informative about 
the absolut metric's sensitivity.

```{r limits, warning=FALSE, message=FALSE }

limit_list <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    lim_rev <- function(x){
        percent <- (max(x) - min(x))/100
        lower <- min(x[x > (max(x) - 10 * percent)]) # smallest value above 0.95 * batch0
        upper <- max(x[x < (min(x) + 10 * percent)])
        ind_low <- which(x == lower)
        ind_up <- which(x == upper)
        batch <- mean_res$batch[c(ind_low, ind_up)]
    }
    batch_lim <- apply(mean_res, 2, lim_rev) %>% set_rownames(c("lower", "upper"))
    batch_lim[,c("kbet", "mm", "pcr", "graph.connectivity")] <- 
      batch_lim[c(2,1), c("kbet", "mm", "pcr", "graph.connectivity")]
    batch_lim
}) %>% set_names(names(mean_tab))

lim_fun <- function(dataset){
    lim_res <- limit_list[[dataset]]
    lim_low <- lim_res["lower",] %>% set_names(colnames(lim_res))
}

limit_tab <- names(limit_list) %>% map(lim_fun) %>% 
    set_names(names(limit_list)) %>% bind_cols() %>% 
    mutate(metric = colnames(limit_list[[1]]))


lim_long <- limit_tab %>% filter(!metric %in% "batch") %>% 
    pivot_longer(-metric, names_to = "dataset", values_to = "batch_limit")


cor_list <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    limit_res <- limit_list[[dataset]]
    cor_fun <- function(metric){
        cor_res <- mean_res  %>%
            filter(batch <= limit_res["upper", metric]) %>%
            select(metric, "batch") %>% cor(method ="spearman") %>% min() %>% abs()
        #%>% filter(batch >= limit_res["lower", metric])
    }
    cor_tab <- map(colnames(mean_res), cor_fun) %>% bind_cols() %>% 
        set_colnames(colnames(mean_res)) 
}) %>% bind_rows() %>% set_rownames(names(mean_tab))

cor_long <- cor_list  %>% mutate(dataset = rownames(.)) %>% select(-batch) %>% 
    pivot_longer(-dataset, names_to = "metric", values_to = "correlation")


cor_list_all <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    limit_res <- limit_list[[dataset]]
    cor_fun <- function(metric){
        cor_res <- mean_res  %>%
            select(metric, "batch") %>% cor(method ="spearman") %>% min() %>% abs()
    }
    cor_tab <- map(colnames(mean_res), cor_fun) %>% bind_cols() %>% 
        set_colnames(colnames(mean_res)) 
}) %>% bind_rows() %>% set_rownames(names(mean_tab))

cor_long_all <- cor_list_all  %>% mutate(dataset = rownames(.)) %>% select(-batch) %>% 
    pivot_longer(-dataset, names_to = "metric", values_to = "correlation")


# # order by metrics type
lim_long$metric <- factor(lim_long$metric,levels = metric_order)
cor_long$metric <- factor(cor_long$metric,levels = metric_order)
cor_long_all$metric <- factor(cor_long_all$metric,levels = metric_order)
```

#### Batch limits
```{r batch lims}
ggplot(lim_long, aes(x = metric , y = batch_limit, fill = dataset)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(lim_long$metric)))], alpha = 0.5) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols) + 
    theme_bw() +
    theme(axis.text.x = element_text(face="bold", size=10, angle=45))
```

#### Relative strength correlation
```{r strength correlation}
ggplot(cor_long, aes(x = metric , y = correlation, fill = dataset)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(cor_long$metric)))], alpha = 0.5) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols_data) + 
    theme_bw() +
    theme(axis.text.x = element_text(face="bold", size=10, angle=45))
```

#### Total strength correlation
```{r strength correlation total}
ggplot(cor_long_all, aes(x = metric , y = correlation, fill = dataset)) + 
    geom_boxplot(fill = cols[1:length(levels(as.factor(cor_long$metric)))], alpha = 0.5) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) +
    scale_fill_manual(values=cols_data) + 
    theme_bw() +
    theme(axis.text.x = element_text(face="bold", size=10, angle=45))
```


#### Plot trade-off between strength correlation and detection limit
```{r relationship}
#join datasets
trade_long <- full_join(cor_long_all, lim_long, by = c("dataset", "metric"))
trade_sorted <- trade_long %>% group_by(metric) %>% 
  summarise(mean_correlation = mean(correlation),
            mean_limit = mean(batch_limit),
            correlation_sd = sd(correlation),
            batch_limit_sd = sd(batch_limit))


lev_metric_level <- c(cell = c("cms.bmin", "cms.batch_id","cms.kmin",
                               "isi", "wisi", "lisi", "entropy", "mm"),
                      celltype = c("kbet", "asw", "graph.connectivity"),
                      global = "pcr")
  
trade_sorted$metric_level <- fct_collapse(factor(trade_sorted$metric), 
                                          cell = c("cms.bmin", "cms.batch_id", 
                                                   "cms.kmin", "isi", "wisi", 
                                                   "lisi", "entropy", "mm"), 
                                          celltype = c("kbet", "asw", "sw", 
                                                       "casw", "graph.connectivity"), 
                                          global = "pcr")

ggplot(trade_sorted, aes(x=mean_limit, y=mean_correlation, color=metric)) +
    geom_pointrange(aes(ymin=mean_correlation - correlation_sd, 
                        ymax=mean_correlation + correlation_sd)) +
    geom_pointrange(aes(xmin=mean_limit - batch_limit_sd, 
                        xmax=mean_limit + batch_limit_sd)) +
    scale_colour_manual(values=cols) +
    ggtitle("Trade off: batch detection limit - correlation") +
    theme_ipsum()
```

#### Trade off top metrics
```{r trade off top metrics}
trade_top <- trade_sorted %>% filter(mean_limit < 0.25, mean_correlation > 0.9)

ggplot(trade_top, aes(x=mean_limit, y=mean_correlation, color=metric)) +
    geom_pointrange(aes(ymin=mean_correlation - correlation_sd, 
                        ymax=mean_correlation + correlation_sd)) +
    geom_pointrange(aes(xmin=mean_limit - batch_limit_sd, 
                        xmax=mean_limit + batch_limit_sd)) +
    scale_colour_manual(values = cols) +
    ggtitle("Trade off: batch detection limit - correlation") +
    theme_ipsum()
```


#### Trade-off mean_only
```{r relationship points}
#join datasets
ggplot(trade_sorted, aes(x=mean_limit, y=mean_correlation, color=metric, shape = metric_level)) +
    geom_point(size = 4.5, color = "black", ) +
    geom_point(size = 4) +
    scale_colour_manual(values=cols) +
    ggtitle("Trade off: batch detection limit - correlation") +
    theme_ipsum()
```


### Correlation between metrics

Check mean correlation
```{r cor, fig.width=9, fig.height=9}

#scale means
scale_m <- lapply(names(mean_tab), function(dataset){
    mean_res <- mean_tab[[dataset]]
    scale_res <- mean_res %>% select(-batch) %>% scale() %>% as.data.frame()
}) %>% set_names(names(mean_tab))

cor_met <- lapply(names(scale_m), function(dataset){
    scal_res <- scale_m[[dataset]]
    cor_scal <- cor(scal_res, use = "complete.obs", method = "spearman") 
})

cor_met_mean <- Reduce(`+`, cor_met)/length(cor_met)

corrplot(cor_met_mean, 
         type="upper", 
         order="original",
         hclust.method = "complete",
         col=brewer.pal(n=8, name="PuOr"),
         addgrid.col = NA,
         addCoef.col = "black",
         diag = FALSE)

#save correlation
saveRDS(cor_met_mean, out_path_cor)
saveRDS(trade_sorted, out_path_res)
```





