---
title: "Metrics Scaling random"
author: "Almut Lütge"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  data:
  meta:
  metrics:
  out_cor:
  fig_res:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: show
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---


```{r dataset name, include=FALSE}
param <- readRDS(params$meta)
dataset_name <- param[["dataset_name"]]

```

## `r paste(dataset_name)`

[Back to home](index.html)

```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(scater)
  library(CellMixS)
  library(purrr)
  library(tidyr)
  library(dplyr)
  library(gridExtra)
  library(scran)
  library(cowplot)
  library(jcolors)
  library(ggpubr)
  library(hrbrthemes)
  library(magrittr)
  library(viridis)
  library(scales)
  library(corrplot)
  library(RColorBrewer)
})

options(bitmapType='cairo')
```


### Dataset and metrics
```{r data}
sce <- readRDS(params$data)
metrics <- params$metrics
metrics

out_path_cor <- params$out_cor
out_path_fig <- params$fig_res

dim_red_nam <- ifelse("TSNE" %in% reducedDimNames(sce), "TSNE", "tsne")

cols_data <-c(c(jcolors('pal6'),jcolors('pal8'))[c(1,8,14,5,2:4,6,7,9:13,15:20)],jcolors('pal4'))
names(cols_data) <- c()

cols <-c(c(jcolors('pal6'),jcolors('pal8'), jcolors('pal7'))[c(1,12,18,25,27,2,4,7,3,6,8,14,9,20)],jcolors('pal4'))

```

## Visualize metric scores

### Effects on score distribution - histograms {.tabset}
How does the metric scores change across batches
```{r vi hist, warning=FALSE, message=FALSE, echo=FALSE}
metrics <- c(names(colData(sce))[names(colData(sce)) %>% grepl("\\_batch0", .)], names(colData(sce))[names(colData(sce)) %>% grepl("\\.batch0", .)])
metrics <- gsub("_batch0", "", metrics)
metrics <- gsub(".batch0", "", metrics)
metrics[which(metrics %in% "cms")] <- "cms.batch"
metrics_raw <- metrics[!grepl("smooth", metrics)]
metrics_raw <-  metrics_raw[!grepl("casw", metrics_raw)]
metrics_raw <-  metrics_raw[!grepl("^sw", metrics_raw)]

#### ---------- Order by metric type ----------------------------------####### 
#(manual needs to be adjusted if new metrics are added)
cms_ind <- grep("cms", metrics_raw)
lisi_ind <- grep("isi", metrics_raw)
ent_ind <- grep("entropy", metrics_raw)
mm_ind <- grep("mm", metrics_raw)
asw_ind <- grep("^asw", metrics_raw)
kbet_ind <- grep("kbet", metrics_raw)
graph_ind <- grep("graph", metrics_raw)
pcr_ind <- grep("pcr", metrics_raw)
metric_order <- metrics_raw[c(cms_ind, lisi_ind[c(1,3,2)], ent_ind, mm_ind, 
                                       kbet_ind, asw_ind, graph_ind, pcr_ind)]
names(cols) <- metric_order
#cols <- cols[names(res_by_met[[1]])]
####--------------------------------------------------------------------########



template <- c(
    "#### {{nm}}\n",
    "```{r hist{{nm}}, echo = FALSE}\n",
    "p <- visHist(sce, metric = '{{nm}}', n_col = 3)\n",
    "p",
    "saveRDS(p, paste0(out_path_fig, '_hist_{{nm}}.rds'))",
    "```\n",
    "\n"
  )

plots <- lapply(metrics_raw, 
  function(nm) knitr::knit_expand(text = template)
)
```

`r knitr::knit(text = unlist(plots))`

### Effects on score distribution - Range plots {.tabset}
How does the metric scores change across batches
```{r vi range, warning=FALSE, message=FALSE, echo=FALSE}

template_range <- c(
    "#### {{nm}}\n",
    "```{r range{{nm}}, echo = FALSE}\n",
    "p <- visIntegration(sce, metric = '{{nm}}', metric_name = '{{nm}}')\n",
    "p",
    "saveRDS(p, paste0(out_path_fig, '_{{nm}}.rds'))",
    "```\n",
    "\n"
  )

plots_range <- lapply(metrics_raw, 
  function(nm) knitr::knit_expand(text = template_range)
)
```
`r knitr::knit(text = unlist(plots_range))`


### Cellspecific score {.tabset}
How does the metric scores change across batches
```{r vi metric, warning=FALSE, message=FALSE, echo=FALSE}

template_tsne <- c(
    "#### {{nm}}\n",
    "```{r tsne{{nm}}, echo = FALSE, fig.height=10,fig.width=10}\n",
    "metric_all_b <- as_tibble(colData(sce)) %>% select(starts_with('{{nm}}')) %>% colnames()",
    "p <- do.call(plot_grid, c(lapply(metric_all_b, function(metric_b){
        visMetric(sce, metric_var = metric_b, dim_red = dim_red_nam) + ggtitle('') +
                guides(color=guide_legend(title='{{nm}}'))
    }), ncol = 3))",
    "p\n",
    "```\n",
    "\n"
  )

plots_tsne <- lapply(metrics, 
  function(nm, dim_red_nam = dim_red_nam) knitr::knit_expand(text = template_tsne)
)
```

`r knitr::knit(text = unlist(plots_tsne))`


## Mean score trend {.tabset}
```{r mean score}

mean_score <- function(score){
  score_all_b <- as_tibble(colData(sce)) %>% select(starts_with(score)) %>% 
    colMeans(na.rm = TRUE)
  mean_tab <- data.frame(score = score_all_b,
                         "randomness" = seq.int(from = 0, to = 100, by = 10))
  #get colour index
  col_ind <- which(metrics_raw %in% score)
  p <- ggplot(mean_tab, aes(x=randomness, y=score)) +
    geom_point(size=4, color=cols[score]) + 
    geom_smooth(method = lm, se = FALSE, color = "black") +
    labs(title=score, x="percent mixed label", y = paste0("average ", score)) +
    scale_color_manual(values=cols) + theme_classic() + stat_cor(method = "pearson")
  plot(p)
}

template_cor <- c(
    "### {{nm}}\n",
    "```{r cor{{nm}}, echo = FALSE}\n",
    "mean_score('{{nm}}')\n",
    "```\n",
    "\n"
  )

plots_cor <- lapply(metrics_raw, 
  function(nm) knitr::knit_expand(text = template_cor)
)
```

`r knitr::knit(text = unlist(plots_cor))`

## Summary/Comparison

```{r sum comp}
#correlation coeff
scaling_sum <- lapply(metrics_raw, function(score){
  score_all_b <- as_tibble(colData(sce)) %>% select(starts_with(score)) %>% 
    colMeans(na.rm = TRUE)
  mean_tab <- data.frame(score = score_all_b,
                         "randomness" = seq.int(from = 0, to = 100, by = 10))
  r <- cor(mean_tab$score, mean_tab$randomness, method = "spearman")
  r_na <- ifelse(is.na(r), 0, r) 
  }) %>% set_names(metrics_raw) %>% bind_cols() %>% t() %>%
  set_colnames("R") %>% as_tibble(.) %>% mutate("metric" = metrics_raw)


#"effect size"
effect_sum <- lapply(metrics_raw, function(score){
  score_all_b <- as_tibble(colData(sce)) %>% select(starts_with(score)) %>% 
    colMeans(na.rm = TRUE)
  effect_tab <- data.frame(score = score_all_b,
                         "percent_score" = score_all_b/score_all_b[11])
  effect_tab_sum <- data.frame("min" = ifelse(score_all_b[1] < score_all_b[11], 
                                              effect_tab[1, "percent_score"],
                                               1/score_all_b[1]), 
                               "range" = effect_tab[11,"score"] - effect_tab[1,"score"])
  }) %>% set_names(metrics_raw) %>% bind_rows() 

scaling_sum <- cbind(scaling_sum, effect_sum)


```

# Combined results
```{r}
#scale means
mean_res <- lapply(metrics_raw, function(score){
  score_all_b <- as_tibble(colData(sce)) %>% select(starts_with(score)) %>% 
    colMeans(na.rm = TRUE)
}) %>% set_names(metrics_raw) %>% bind_cols() %>% 
    mutate(randomness = seq.int(from = 0, to = 100, by = 10))

#correlation
cor_metrics <- cor(mean_res, method = "spearman")

dir_ind <- which(!colnames(mean_res) %in% c("randomness", "mm", "kbet", "pcr"))
mean_res[, dir_ind] <- mean_res[, dir_ind] * -1
scale_res <-  mean_res %>% select(-randomness) %>% scale() %>% as.data.frame()
scale_res[is.na(scale_res)] <- 0  
scale_res$randomness <- mean_res$randomness

mean_long <- scale_res %>% pivot_longer(-randomness, names_to = "metric",
                                           values_to = "scaled_result")

mean_long$metric <- factor(mean_long$metric,levels = metric_order)

ggplot(mean_long, aes(x = randomness, y = scaled_result)) +
    geom_line(aes(color = metric)) + 
    geom_point(aes(color = metric)) +
    theme_bw() +
    scale_color_manual(values = cols[levels(as.factor(mean_long$metric))]) 


```



## Separated plots
```{r spagetti separated,  fig.width = 8, fig.height = 7}

scale_long <- scale_res %>% pivot_longer(-randomness, names_to = "metric",
                                           values_to = "scaled_score")

scale_long$metric <- factor(scale_long$metric,levels = metric_order)
cols_rep <- rep(cols[levels(as.factor(scale_long$metric))], each = nrow(mean_res))
scale_long$metric <- recode(scale_long$metric, cms.kmin80 = "cms_kmin", cms.bmin80 = "cms_bmin",
       cms.batch = "cms_default", graph_connectivity = "graph", kbet = "kBet")

label <- data.frame(
  scaled_score = c(rep(1.5, 1)), 
  randomness = c(rep(65, 12)), 
  metric = levels(scale_long$metric),
  label = c(paste0("R = ", signif(abs(cor_metrics[metric_order,"randomness"]), digits = 2)))
)

scale_long <- scale_long %>% mutate(metric2 = metric)
p <- ggplot(scale_long, aes(x = randomness, y = scaled_score)) +
      geom_line(data=scale_long %>% dplyr::select(-metric), aes(group=metric2),
                 color="grey", size=0.5, alpha=0.5) + 
      geom_line( aes(color=metric), color=cols_rep, size=1.2 ) + 
      scale_color_viridis(discrete = TRUE) +
      theme_ipsum(base_family = 'Helvetica') +
      theme(
        legend.position="none",
        plot.title = element_text(size=14),
        panel.grid.minor = element_blank()
      ) +
      geom_label(data = label, aes(label = label), size = 6) + 
      ggtitle("Gradual increased randomness") +
      facet_wrap(~metric) 
p

saveRDS(p, paste0(out_path_fig, "_all.rds"))

```

# Correlation of metrices
```{r correlation, fig.width=9, fig.height=9}

corrplot(cor_metrics, 
         type="upper", 
         order="original",
         hclust.method = "complete",
         col=brewer.pal(n=8, name="PuOr"),
         addgrid.col = NA,
         addCoef.col = "black",
         diag = FALSE)

#save correlation
saveRDS(cor_metrics, out_path_cor)

```



```{r}
sessionInfo()
```

