---
title: "Figures benchmark"
author: "Almut Lütge"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  random_tsne:
  random_res_all:
  random_pbmc:
  random_lisi:
  random_ent:
  random_cms:
  random_mm:
  random_pcr:
  random_kbet:
  out_path:
  sca_limit:
  sca_cor:
  sim_data:
  sca_trade_off:
  unb_moderate:
  unb_limit:
  cor:
  cor_legend:
  summary:
  cor_char:
  h_size:
  h_type:
  h_ct_spec:
  lfc_dist:
  tsne_type:
  rel_cells:
  rel_genes:
  sim_scal_dist:
  tern:
  tern_nopan:
  h_gi:
  gi_legend:
  den1:
  den2:
  met_char:
  tern_sim:
  den3:
  den4:
  den5:
  sim1:
  sim2:
  sim3:
  time:
  mem:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: show
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Benchmark figures mixing metrics

```{r libs}
suppressPackageStartupMessages({
  library(cowplot)
  library(hrbrthemes)
  library(ggplot2)
  library(jcolors)
  library(ComplexHeatmap)
  library(dplyr)
})

cols_data <-c(c(jcolors('pal6'),jcolors('pal8'))[c(1,8,14,5,2:4,6,7,9:13,15:20)],jcolors('pal4'))
names(cols_data) <- c()
```

Get figure objects
```{r data}

# random mixing
random_tsne <- readRDS(params$random_tsne)
random_res_all <- readRDS(params$random_res_all)
random_pbmc <- readRDS(params$random_pbmc)
random_cms <- readRDS(params$random_cms)
random_lisi <- readRDS(params$random_lisi)
random_ent <- readRDS(params$random_ent)
random_mm <- readRDS(params$random_mm)
random_pcr <- readRDS(params$random_pcr)
random_kbet <- readRDS(params$random_kbet)

# scaling and batch limits simulation
limit_scaling <- readRDS(params$sca_limit)
correlation_scaling <- readRDS(params$sca_cor)
trade_off <- readRDS(params$sca_trade_off)
sim_data <- readRDS(params$sim_data)
sim_scal_dist <- readRDS(params$sim_scal_dist)

# unbalanced mixing
unb_moderate <- readRDS(params$unb_moderate)
unb_limit <- readRDS(params$unb_limit)

# batch characteristics
cor_char <- readRDS(params$cor_char)
h_size <- readRDS(params$h_size)
h_type <- readRDS(params$h_type)
h_ct_spec <- readRDS(params$h_ct_spec)
lfc_dist <- readRDS(params$lfc_dist)
tsne_type <- readRDS(params$tsne_type)
rel_cells <- readRDS(params$rel_cells)
rel_genes <- readRDS(params$rel_genes)
gi_legend <- readRDS(params$gi_legend)
h_gi <- readRDS(params$h_gi)
tern <- readRDS(params$tern)
tern_nopan <- readRDS(params$tern_nopan)
den_1 <- readRDS(params$den1)
den_2 <- readRDS(params$den2)
met_char <- readRDS(params$met_char)

# correlation (and summary?)
cor <- readRDS(params$cor)
cor_legend <- readRDS(params$cor_legend)
summary <- readRDS(params$summary)

# general
out_path <- params$out_path


#Simulation
tern_sim <- readRDS(params$tern_sim)
den_3 <- readRDS(params$den3)
den_4 <- readRDS(params$den4)
den_5 <- readRDS(params$den5)
sim_1 <- readRDS(params$sim1)
sim_2 <- readRDS(params$sim2)
sim_3 <- readRDS(params$sim3)


# time and memory
time <- readRDS(params$time)
mem <- readRDS(params$mem)


#themes
theme_def <- theme_ipsum(base_family = 'Helvetica',
                    strip_text_face = "bold",
                    axis_title_size = 25,
                    strip_text_size = 22,
                    base_size = 25) +
        theme(
            legend.position="top",
            plot.title = element_text(size=26),
            plot.subtitle = element_text(size=20),
            panel.grid.major = element_blank(),
            axis.text.x = element_blank(),
            axis.text.y = element_blank(),
            panel.spacing = unit(1, "lines")
        )

theme_axis <- theme_ipsum(base_family = 'Helvetica',
                    strip_text_face = "bold",
                    axis_title_size = 25,
                    strip_text_size = 22,
                    plot_margin = margin(5, 5, 5, 5),
                    base_size = 25) +
        theme(
            legend.position="top",
            plot.title = element_text(size=26),
            axis.text.x = element_text(size=16),
            panel.grid.major = element_blank(),
            panel.spacing = unit(1, "lines")
        )

theme_axis_long <- theme_ipsum(base_family = 'Helvetica',
                    strip_text_face = "bold",
                    axis_title_size = 25,
                    strip_text_size = 22,
                    base_size = 25) +
        theme(
            legend.position="top",
            plot.title = element_text(size=26),
            panel.grid.major = element_blank(),
            axis.text.x = element_text(size=16, angle=45, hjust = 1),
            panel.spacing = unit(1, "lines")
        )

theme_summary <- theme_ipsum(base_family = 'Helvetica',
                    strip_text_face = "bold",
                    axis_title_size = 24,
                    strip_text_size = 22,
                    base_size = 22) +
        theme(
            legend.position="top",
            plot.title = element_text(size=26),
            axis.text.x = element_text(size=18),
            panel.spacing = unit(1, "lines")
        )
```

### Figure Benchmark random

```{r fig random, fig.width = 16, fig.height=18}
p1 <- random_tsne
batch_int <- c("batch0", "batch30", "batch60", "batch100")
p1_1 <- p1 %+% p1$data[p1$data$randomness %in% batch_int,]
p1_1$layers[[1]]$aes_params$size <- 1
p1_1$layers[[1]]$aes_params$alpha <- 0.5 
p1_1 <- p1_1 + guides(colour = guide_legend(override.aes = list(size=7))) + 
  scale_colour_manual(values = c("#1D2645FF", "#E75B64FF")) + 
  theme_def
p1_1$facet$params$ncol <- 4

p2 <- random_res_all + 
  labs(title = "Mean score with increased number of randomly shuffled labels", 
       subtitle = "Dataset with clearly separated batches (see A)") +
  theme_def +
  xlab("% random batch label") +
  theme(axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18))

p2$data$metric <- recode(p2$data$metric, cms.kmin80 = "cms_kmin", cms.bmin80 = "cms_bmin",
       cms.batch = "cms_default", graph_connectivity = "graph", kbet = "kBet")

p2$data$metric2 <- recode(p2$data$metric2, cms.kmin80 = "cms_kmin", cms.bmin80 = "cms_bmin",
       cms.batch = "cms_default", graph_connectivity = "graph", kbet = "kBet")
p2$facet$params$ncol <- 6

p_random <- plot_grid(p1_1, p2, labels = c('A', 'B'), label_size = 20, ncol = 1,
                      rel_heights = c(1.4, 2))
p_random

ggsave(filename = paste0(out_path, "results_random.pdf"), plot = p_random,
       dpi = 600)
```

Supplementary figure random
```{r supp random, fig.width=12, fig.height= 12}
p3 <- random_pbmc + 
  labs(title = "Mean score with increased number of randomly shuffled labels", 
       subtitle = "Dataset with moderate batch effect") +
  theme_axis

p3$data$metric <- recode(p3$data$metric, cms.kmin80 = "cms_kmin", cms.bmin80 = "cms_bmin",
       cms.batch = "cms_default", graph_connectivity = "graph", kbet = "kBet")

p3$data$metric2 <- recode(p3$data$metric2, cms.kmin80 = "cms_kmin", cms.bmin80 = "cms_bmin",
       cms.batch = "cms_default", graph_connectivity = "graph", kbet = "kBet")

p3 <- plot_grid(p3)
p3

ggsave(filename = paste0(out_path, "results_random_pbmc.pdf"), plot = p3,
       dpi = 600)
```


```{r random supp met, fig.width= 17, fig.height= 20}
p1 <- random_cms + ylab("random batch label [%]") + 
  xlab("cms_default") + 
  labs(title = "Cms score distributions") +  theme_def +
  scale_y_discrete(breaks=c("cms.batch0","cms.batch20","cms.batch40", "cms.batch60",
                            "cms.batch80", "cms.batch100"),
                   labels=c("0", "20", "40", "60", "80", "100")) +
  theme(legend.position = "none",
        axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18))

p2 <- random_lisi + ylab("") + 
  labs(title = "Lisi score distributions") +  theme_def + 
  xlab("lisi") +
  scale_y_discrete(breaks=c("lisi.batch0","lisi.batch20","lisi.batch40", "lisi.batch60",
                            "lisi.batch80", "lisi.batch100"),
                   labels=c("0", "20", "40", "60", "80", "100")) + 
  theme(legend.position = "none",
        axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18))

p3 <- random_ent + ylab("random batch label [%]") + 
  xlab("entropy") +
  labs(title = "Entropy score distributions") +  theme_def + 
  scale_y_discrete(breaks=c("entropy_batch0","entropy_batch20","entropy_batch40", "entropy_batch60",
                            "entropy_batch80", "entropy_batch100"),
                   labels=c("0", "20", "40", "60", "80", "100")) +
  theme(legend.position = "none",
        axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18))

p4 <- random_mm + ylab("") + 
  xlab("mixing metric") +
  labs(title = "Mm score distributions") +  theme_def +
  scale_y_discrete(breaks=c("mm_batch0","mm_batch20","mm_batch40", "mm_batch60",
                            "mm_batch80", "mm_batch100"),
                   labels=c("0", "20", "40", "60", "80", "100")) +
  theme(legend.position = "none",
        axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18))

p5 <- random_pcr + ylab("random batch label [%]") + 
  xlab("pcr") +
  labs(title = "Pcr score distributions") +  theme_def +
  scale_y_discrete(breaks=c("pcr.batch0","pcr.batch20","pcr.batch40", "pcr.batch60",
                            "pcr.batch80", "pcr.batch100"),
                   labels=c("0", "20", "40", "60", "80", "100")) +
  theme(legend.position = "none",
        axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18))

p6 <- random_kbet + ylab("") + 
  xlab("kBet") +
  labs(title = "kBet score distributions") +  theme_def +
  scale_y_discrete(breaks=c("kbet.batch0","kbet.batch20","kbet.batch40", "kbet.batch60",
                            "kbet.batch80", "kbet.batch100"),
                   labels=c("0", "20", "40", "60", "80", "100")) +
  theme(legend.position = "none",
        axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18))

p_ran_met <- plot_grid(p1, p2, p3, p4, p5, p6, 
                       labels = c('A', 'B', 'C', 'D', 'E', 'F'), label_size = 20, 
                       ncol = 2)
p_ran_met

ggsave(filename = paste0(out_path, "results_random_mets.pdf"), plot = p_ran_met,
       dpi = 600)
```



### Figure Scaling and batch limits
```{r scaling and batch limits, fig.width=16, fig.height = 18}
p2 <- limit_scaling + 
  theme_axis_long +
  theme(legend.position = "none") + 
  ggtitle("Batch detection limits")
p2$data$metric <- recode(p2$data$metric, cms.kmin = "cms_kmin", cms.bmin = "cms_bmin",
       cms.batch_id = "cms_default", graph.connectivity = "graph", kbet = "kBet")
p1 <- correlation_scaling + 
  theme_axis_long +
  theme(legend.position = "none") + 
  ggtitle("Scaling")
p1$data$metric <- recode(p1$data$metric, cms.kmin = "cms_kmin", cms.bmin = "cms_bmin",
       cms.batch_id = "cms_default", graph.connectivity = "graph", kbet = "kBet")

  
p3 <- sim_data 
p3$layers[[1]]$aes_params$size <- 1
p3$layers[[1]]$aes_params$alpha <- 0.5 
sim_int <- c("0", "0.6", "1", "2")
p3_1 <- p3 %+% p3$data[p3$data$batch_size %in% sim_int,]
p3_1 <- p3_1 + 
  scale_colour_manual(values = c("#1D2645FF", "#E75B64FF")) + 
  theme_def + theme(legend.position = "bottom") +
  labs(title = "Simulation with incresing batch logFC")

p4 <- trade_off +
  theme_axis_long +
  xlab("mean limit") +
  theme(legend.position = "bottom",
        legend.box = "vertical",
        legend.title = element_text(size = 18),
        legend.text = element_text(size = 16)) 
p4$data$metric <- recode(p4$data$metric, cms.kmin = "cms_kmin", cms.bmin = "cms_bmin",
       cms.batch_id = "cms_default", graph.connectivity = "graph", kbet = "kBet")

#p4 <- plot_grid(p4, "", ncol = 1, rel_heights = c(2,0.1))

p_sub1 <- plot_grid(p3_1, p1, labels = c('A', 'B'), align = "v", 
                   label_size = 20, ncol = 2, rel_heights = c(1,1))

p_sub2 <- plot_grid(p2, p4, labels = c('C', 'D'), label_size = 20,
                   ncol = 2, rel_heights = c(1, 1))


p_scaling <- plot_grid(p_sub1, p_sub2, ncol = 1, rel_heights = c(1, 1))
p_scaling

ggsave(filename = paste0(out_path, "results_scaling.pdf"), plot = p_scaling,
       dpi = 600)
```

#### Supplement figure: Score distributions scaling simulated
```{r supp scaling sim, fig.width=12, fig.height=12}
p_dist <- sim_scal_dist + 
  labs(title = "Mean score by relative batch strength",
       subtitle = "Simulation series of 13 simulations with increasing batch effect.") +
  xlab("batch logFC factor") +
  theme_axis + theme(axis.text.x = element_text(size=22))

p_dist$data$metric <- recode(p_dist$data$metric, cms.kmin = "cms_kmin", cms.bmin = "cms_bmin",
       cms.batch_id = "cms_default", graph.connectivity = "graph", kbet = "kBet")

p_dist$data$metric2 <- recode(p_dist$data$metric2, cms.kmin = "cms_kmin", cms.bmin = "cms_bmin",
       cms.batch_id = "cms_default", graph.connectivity = "graph", kbet = "kBet")

p_dist$layers[[3]]$data$metric <- recode(p_dist$layers[[3]]$data$metric,
                                              cms.kmin = "cms_kmin", 
                                              cms.bmin = "cms_bmin", 
                                              cms.batch_id = "cms_default", 
                                              graph.connectivity = "graph", 
                                              kbet = "kBet")

p_dist$layers[[4]]$data$metric <- recode(p_dist$layers[[3]]$data$metric,
                                              cms.kmin = "cms_kmin", 
                                              cms.bmin = "cms_bmin", 
                                              cms.batch_id = "cms_default", 
                                              graph.connectivity = "graph", 
                                              kbet = "kBet")


p_dist <- plot_grid(p_dist)
p_dist

ggsave(filename = paste0(out_path, "results_sim_scal_dist.pdf"), plot = p_dist,
       dpi = 600)
```



### Figure unbalanced batch effects
```{r unbalanced, fig.width=16, fig.height = 22}
unb_moderate$layers[[3]]$data$metric <- recode(unb_moderate$layers[[3]]$data$metric,
                                              cms.kmin = "cms_kmin", 
                                              cms.bmin = "cms_bmin", 
                                              cms.batch_id = "cms_default", 
                                              graph.connectivity = "graph", 
                                              kbet = "kBet")


p1 <- unb_moderate + 
  theme_axis_long + 
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 22),
        axis.text.x = element_text(size=16)) +
  scale_y_continuous(breaks=c(-0.3, 0, 0.1)) +
  scale_x_continuous(breaks = c(0,0.5, 1), labels = c("0", "50", "100")) +
  xlab("[%] unbalanced") + 
  ggtitle("Metric scores by increasingly imbalanced batch effects",
          subtitle = "Simulated data with a moderate batch effect using csf_patient as reference.")
p1$data$metric <- recode(p1$data$metric, cms.kmin = "cms_kmin", cms.bmin = "cms_bmin",
       cms.batch_id = "cms_default", graph.connectivity = "graph", kbet = "kBet")

# p2 <- unb_limit + 
#   theme_axis_long +
#   ylab('unbalance limit') +
#   theme(legend.position = "top",
#         axis.text.x = element_text(size=20, angle=45, hjust = 1)) + 
#   ggtitle("Imbalance limits of metric score's divergence")
p2 <- unb_limit + 
  theme_axis_long +
  ylab('imbalance limit') +
  theme(legend.position = "top",
        plot.subtitle = element_text(size = 22),
        axis.text.x = element_blank(),
        strip.text.x = element_text(size = 16, angle=45, hjust = 0.2)) + 
  ggtitle("Imbalance limits of metric score's divergence",
          subtitle = "Simulated data with csf_patient as reference.")

# p2$data$metric <- recode(p2$data$metric, cms.kmin = "cms_kmin", cms.bmin = "cms_bmin",
#        cms.batch_id = "cms_default", graph.connectivity = "graph", kbet = "kBet")

p2$layers[[2]]$aes_params$size <- 7
 
p_unbalanced <- plot_grid(p1, "", p2, labels = c('A','', 'B'), label_size = 20, ncol = 1,
                          rel_heights = c(14,0.5,9))
p_unbalanced

ggsave(filename = paste0(out_path, "results_unbalanced.pdf"), plot = p_unbalanced,
       dpi = 600)

```


### Summary and correlation
```{r sum cor, fig.width= 13, fig.height = 12, warning=FALSE, message=FALSE}
p2 <- plot_grid(plotlist = cor)
p2_legend <-  cor_legend
p2_ <- plot_grid("", p2, p2_legend, ncol = 3, rel_widths = c(0.1, 1, 0.3))
p2_title <- ggdraw() + 
  draw_label(
    "Metrics correlation - by benchmark task",
    fontface = 'bold',
    x = 0,
    size = 20,
    hjust = - 0.3,
    vjust = 0.5) +
  theme_def +
  theme(plot.margin = margin(0, 0, 0, 7))
p2_new <- plot_grid(p2_title, p2_, "", ncol = 1, rel_heights = c(0.15, 1, 0.15))
p2_new

p1 <- summary + theme_summary + 
  theme(legend.position = "top")

p1$data$metric <- recode(p1$data$metric, graph.connectivity = "graph", kbet = "kBet")

p1$layers[[1]]$aes_params$size <- 16 

p1_new <- plot_grid("", p1, "", ncol = 1, labels = c('A','', ''), 
                    label_size = 20, rel_heights = c(0.01, 1, 0.01))

p_summary <- plot_grid(p1) 
p_summary
p_cor <- plot_grid(p2_new)
p_cor

ggsave(filename = paste0(out_path, "results_summary.pdf"), plot = p_summary, dpi = 600)
ggsave(filename = paste0(out_path, "results_correlation.pdf"), plot = p_cor, dpi = 600)
```

### Figure characterization supplement
```{r characterization, fig.width=12, fig.height = 15}

## Plot characteristics
p_size <- grid.grabExpr(draw(h_size))
p_ct <- grid.grabExpr(draw(h_ct_spec))
pct_title <- ggdraw() + 
  draw_label("Batch characteristics",
    fontface = 'bold', x = 0, size = 20, hjust = - 0.3, vjust = 0.5) +
  theme(plot.margin = margin(0, 0, 0, 7))
p_ct1 <- plot_grid(pct_title, p_ct, "", labels = c('', '', ''), label_size = 20, 
                   ncol = 1, rel_heights = c(2, 10, 1))

p_lfc <- grid.grabExpr(draw(lfc_dist))
plfc_title <- ggdraw() + 
  draw_label("Batch logFC distribution across cell-types",
    fontface = 'bold', x = 0, size = 20, hjust = - 0.3, vjust = 0.5) +
  theme(plot.margin = margin(0, 0, 0, 7))
p_lfc1 <- plot_grid(plfc_title, p_lfc, ncol = 1, rel_heights = c(1, 10))
p_lfc2 <- plot_grid('', p_lfc1, '', labels = c('B', '', ''), label_size = 20, 
                   ncol = 3, rel_widths = c(1, 10, 0.5))

p_char <- plot_grid(p_ct1, p_lfc2, ncol = 1, rel_heights = c(1.5, 1))
p_char


### Plot batch types
p_type <- grid.grabExpr(draw(h_type))
ptype_title <- ggdraw() + 
  draw_label("Batch types: Bias after adjustment",
    fontface = 'bold', x = 0, size = 20, hjust = - 0.3, vjust = 0.5) +
  theme(plot.margin = margin(0, 0, 0, 7))
p_type1 <- plot_grid(ptype_title, p_type, "", labels = c('B', '', ''), 
                     label_size = 20, ncol = 1, rel_heights = c(2, 10, 1))
p_tsne <- tsne_type + theme_def +
  theme(plot.title = element_text(size = 20))
p_tsne <- plot_grid('', p_tsne, '', labels = c('A', '', ''), 
                    label_size = 20, ncol = 3, rel_widths = c(0.4,10, 0.7))

p3 <- plot_grid(p_tsne, p_type1, ncol = 1, rel_heights = c(1.1, 1) )
p3

### Save plots
ggsave(filename = paste0(out_path, "results_type.pdf"), plot = p3, dpi = 600)
ggsave(filename = paste0(out_path, "results_char.pdf"), plot = p_char, dpi = 600)
```


### Figure Characterization short supplement 
```{r char short, fig.width=12, fig.height = 8}
ggsave(filename = paste0(out_path, "results_char_short.pdf"), plot = p_ct1, dpi = 600)
```



Results summary metrics and batch characteristics
```{r char split 1, fig.width=14, fig.height = 14}
# p1_1 <- cor_char + theme_axis_long + 
#   theme(legend.position = "right",
#         axis.text.x = element_text(size=22, angle=45, hjust = 1))
# 
# p_char_metrics <- plot_grid('', p1_1, ncol = 2,
#                 rel_widths = c(0.5,10))
#   
# p_char_metrics
# 
# ggsave(filename = paste0(out_path, "results_char_metrics.pdf"), plot = p1_1,
#        dpi = 600)
met_char$data$metric <- recode(met_char$data$metric, cms.kmin80 = "cms_kmin", cms.bmin80 = "cms_bmin",
       cms = "cms_default", graph_connectivity = "graph", kbet = "kBet")


p_task1 <- met_char + 
  theme_axis +
  xlab('mean PVE-Batch') +
  ylab('scaled metric score') +
  guides(linetype=guide_legend(keywidth = 3, keyheight = 1)) +
  scale_linetype_discrete("characteristics", labels = c("DE genes", "PVE-Batch")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(size=25)) + 
  ggtitle("Metric scores and batch characteristics across datasets")

p_task1

ggsave(filename = paste0(out_path, "results_char_metrics.pdf"), plot = p_task1)
```


### Figure characterization varPart
```{r characterization tern, fig.width=15, fig.height = 20}
library(ggtern)
## Plot characteristics
p_gi <- grid.grabExpr(draw(h_gi, annotation_legend_list = gi_legend))
p_gi1 <- plot_grid('', p_gi, '', labels = c('A', '', ''), label_size = 20, 
                   ncol = 3, rel_widths = c(0.1, 2, 0.1))
p_tern <- ggplotGrob(tern)
den1 <- grid.grabExpr(draw(den_1, heatmap_legend_side = "bottom", annotation_legend_side = "bottom"))
den2 <- grid.grabExpr(draw(den_2, heatmap_legend_side = "bottom", annotation_legend_side = "bottom"))

p_tern1 <- plot_grid(p_tern, labels = c('A'), label_size = 20, ncol = 1)

p_1 <- plot_grid('', den1, '', den2, '', ncol = 5, labels = c('B','', 'C', '', ''), label_size = 20, 
                 rel_widths = c(0.2, 2, 0.1, 2, 0.2))

plfc_title <- ggdraw() + 
  draw_label("Batch logFC distribution across cell types",
    fontface = 'bold', x = 0, size = 22, hjust = - 0.2, vjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 0))

p_den1 <- plot_grid(plfc_title, p_1, ncol = 1, labels = c('', ''), label_size = 20, 
                    rel_heights = c(1, 10))

p_tern_den <-  plot_grid(p_tern1,'', p_den1, ncol = 1, rel_heights = c(1.7, 0.04, 0.8))

# p_char2 <-  plot_grid(p_gi1, p_tern_den, ncol = 1, rel_heights = c(0.7, 3), 
#                       labels = c('', ''), label_size = 20)

### Save plots
ggsave(filename = paste0(out_path, "results_char_tern.pdf"), plot = p_tern_den, dpi = 600)
```


### Supplementary figure(s) Simulation
```{r sim tern, fig.width=18, fig.height = 20}
tern1_title <- ggdraw() + 
  draw_label("Real data",
    fontface = 'bold', x = 0, size = 30, hjust = - 0.5, vjust = 0.5) +
  theme(plot.margin = margin(0, 0, 0, 0))

tern2_title <- ggdraw() + 
  draw_label("Simulated data",
    fontface = 'bold', x = 0, size = 30, hjust = - 0.5, vjust = 0.5) +
  theme(plot.margin = margin(0, 0, 0, 0))

p_tern_sim <- ggplotGrob(tern_sim)
p_tern_nopan <- ggplotGrob(tern_nopan)
p_tern1 <- plot_grid(tern1_title, p_tern_nopan, labels = c('A', ''), label_size = 20, ncol = 1, rel_heights = c(0.2,10))
p_tern2 <- plot_grid(tern2_title, p_tern_sim, labels = c('B', ''), label_size = 20, ncol = 1, rel_heights = c(0.2,10))

p_tern_com <-  plot_grid(p_tern1, "", p_tern2, ncol = 3, rel_widths = c(1, 0.1, 1))

### Save plots
ggsave(filename = paste0(out_path, "results_sim_tern.pdf"), plot = p_tern_com, dpi = 600)
```


```{r sim den, fig.width=16, fig.height = 18}
den1_title <- ggdraw() + 
  draw_label("Real data",
    fontface = 'bold', x = 0, size = 30, hjust = -0.5, vjust = 0.5) +
  theme(plot.margin = margin(0, 0, 0, 0))

den2_title <- ggdraw() + 
  draw_label("Simulated data",
    fontface = 'bold', x = 0, size = 30, hjust = -0.5, vjust = 0.5) +
  theme(plot.margin = margin(0, 0, 0, 0))

den1 <- grid.grabExpr(draw(den_3))
den2 <- grid.grabExpr(draw(den_4))
den3 <- grid.grabExpr(draw(den_5))

sim1 <- grid.grabExpr(draw(sim_1))
sim2 <- grid.grabExpr(draw(sim_2))
sim3 <- grid.grabExpr(draw(sim_3))

p_den <- plot_grid(den1_title, den1, den2, den3, labels = c('', 'A', 'C', 'E'), label_size = 20, ncol = 1, rel_heights = c(0.2,2,2,2,2))
p_sim <- plot_grid(den2_title, sim1, sim2, sim3, labels = c('', 'B', 'D', 'F'), label_size = 20, ncol = 1, rel_heights = c(0.2,2,2,2,2))

p_den_com <-  plot_grid(p_den, p_sim, ncol = 2, rel_widths = c(1, 1))

### Save plots
ggsave(filename = paste0(out_path, "results_sim_den.pdf"), plot = p_den_com, dpi = 600)
```



## Supplemental figure: Correlation RVEB and number of cells resp. genes
```{r supp ext factor, fig.width=12, fig.height = 18}
p_cells <- rel_cells + 
  ggtitle("Number of cells")
p_genes <- rel_genes +
  ggtitle("Number of genes")

p_rel <- plot_grid(p_cells, p_genes, ncol = 2,
                rel_widths = c(1,1))
  
p_rel

ggsave(filename = paste0(out_path, "results_char_rel_corr_ext.pdf"), plot = p_rel,
       dpi = 600)
```


## Time and memory 
```{r time mem, fig.width=13, fig.height = 10}
p_mem <- mem + theme_ipsum(base_family = 'Helvetica',
                    strip_text_face = "bold",
                    axis_title_size = 25,
                    strip_text_size = 22,
                    plot_margin = margin(5, 5, 5, 5),
                    base_size = 23) +
        theme(
            legend.position="none",
            plot.title = element_text(size=26),
            panel.grid.major = element_blank(),
            panel.spacing = unit(1, "lines")
        ) + ggtitle("RSS")
p_time <- time + theme_ipsum(base_family = 'Helvetica',
                    strip_text_face = "bold",
                    axis_title_size = 25,
                    strip_text_size = 18,
                    base_size = 25) +
        theme(
            legend.position="bottom",
            plot.title = element_text(size=26),
            axis.text.x = element_text(size=25),
            axis.text.y = element_text(size=25)
        )



p_time_mem <- plot_grid(p_time, p_mem, ncol = 1,
                rel_heights = c(1,0.9), labels = c('A', 'B'), label_size = 20)


p_time2 <- plot_grid(p_time)

ggsave(filename = paste0(out_path, "results_time_mem.pdf"), plot = p_time2,
       dpi = 600)

```



